---
title: "Functional programming for ecological data"
date: 2026-07-21
categories: [functional programming, purrr, data.table, dplyr]
description: "Twenty ~10-minute quests — use understanding to verify AI-generated R, from base functions to purrr, furrr, and data.table."
toc: true
toc-depth: 2
format:
  html:
    code-fold: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4.5,
  dpi = 120
)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(data.table)
  library(ggplot2)
  library(plotly)
})
```

## Use AI to defeat AI

Large language models are excellent at **pattern-matching** R code they have seen before. They are weaker at knowing *your* survey design, which column is the key, or whether `map()` should have been `map2()`. If you only copy-paste, the bug ships with the manuscript.

This track trains the skill AI cannot give you: **reading and owning the code**. Each quest is ~10 minutes — short segments improve retention in technical material ([Mayer, 2009](https://doi.org/10.1016/B978-0-12-374921-5.00009-3)). We use a predictable rhythm from classroom research ([Rosenshine, 2012](https://doi.org/10.1177/0031721712452808)):

1. **Tell you** what you will accomplish and which packages you need.
2. **Explain** any new function *before* it appears in a chunk.
3. **Show** annotated code — one comment per line so you can narrate it aloud.
4. **Apply** where this shows up in marine ecology workflows (and what to ask AI to do next).

When AI suggests a pipeline, you will know enough to say *“show me the list structure first”* or *“that needs `safely()` for my QC flags.”* That is the win.

**Season 1:** 20 quests · same reprex dataset throughout · mark quests done in your script header for a cheap **retrieval** boost ([Roediger & Butler, 2011](https://doi.org/10.1016/j.jml.2010.09.004)).

::: {.quest-hub}
<p class="quest-hub__title">Season 1 · Functional programming</p>
<p class="quest-hub__hint">You vs the autocomplete: build recipes you trust, then let AI draft the boring lines. Roadmap for quests 21–100 at the bottom.</p>

::: {.quest-levels}
::: {.quest-level}
**Level 1**
Scout · base R · quests 1–5
:::
::: {.quest-level}
**Level 2**
Wrangler · dplyr · quests 6–10
:::
::: {.quest-level}
**Level 3**
Mapper · purrr · quests 11–16
:::
::: {.quest-level}
**Level 4**
Parallel diver · furrr & data.table · quests 17–20
:::
:::
:::

## Reprex: reef transect counts

Before Quest 1, create an RStudio project and run this once. **`tibble()`** builds a modern data frame; **`set.seed()`** fixes randomness so your reprex matches ours when you ask AI for help.

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
`base R`, **`tibble`** (install via `install.packages("tibble")` if needed — it ships with the tidyverse).
:::
::: {.quest-learn__row}
**By the end you will**
Have a 60-row table of simulated belt-transect counts you can paste into any forum or prompt.
:::
:::

```{r reprex-data}
set.seed(427) # fix randomness so counts match this tutorial exactly

reef_transects <- tibble::tibble( # build a tidy data frame
  transect_id = rep(sprintf("T%02d", 1:12), each = 5), # 12 transects, 5 species rows each
  site        = rep(rep(c("North", "South", "East"), length.out = 12), each = 5), # reef site label
  species     = rep( # five fish names repeated across transects
    c("Parrotfish", "Surgeonfish", "Grouper", "Snapper", "Damselfish"),
    times = 12
  ),
  count       = rpois(60, lambda = c(8, 12, 3, 5, 20)), # Poisson counts — common for count data
  depth_m     = rep(round(runif(12, 2, 15), 1), each = 5) # depth per transect, repeated per species row
)

reef_transects # print to confirm structure
```

::: {.callout-tip appearance="simple"}
## RStudio habit

Save as `00-reprex-reef.R`. Run line-by-line (`Cmd/Ctrl + Enter`). When AI gives you code, paste into a **blank script** first — never straight into your only copy of the analysis.
:::

---

::: {.quest-step}
::: {.quest-step__meta}
Quest 1 · ~10 min · Level 1 — Scout
:::

### What is a function, really? {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
`base R` only.
:::
::: {.quest-learn__row}
**By the end you will**
Define a small function, call it with named arguments, and inspect a built-in function object — so AI-generated `function(...)` blocks are not magic.
:::
::: {.quest-learn__row}
**Apply it later**
Wrap any ecology calculation you repeat (CPUE, biomass, richness) in a named function, then ask AI to *extend* your function instead of inventing a new one from scratch.
:::
:::

**What you'll learn:** A function is a **named recipe**. **`function(arg1, arg2) { ... }`** creates it; **`mean`**, **`sum`**, **`lm`** are functions you already call. Printing **`mean`** shows **formals** (arguments), **body** (code), and **environment** — the three parts every function has ([Wickham, 2014](https://adv-r.hadley.nz/function-objects.html)).

```{r quest-1}
mean # show the function object (formals, body, environment)

fish_biomass <- function(count, mean_weight_kg) { # define inputs: counts and average weight
  count * mean_weight_kg # body: multiply vector count by scalar or vector weight
} # implicit return of last evaluated line

fish_biomass( # call our recipe on one row of data
  count = reef_transects$count[1], # first count in the table
  mean_weight_kg = 0.4 # example weight in kg for that species group
)
```

**Check yourself:** Write **`richness()`** that takes a data frame and returns **`length(unique(species_column))`**. Test on **`filter(reef_transects, site == "North")`**. If AI writes it for you, read each line aloud before running.
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 2 · ~10 min
:::

### Vectorised thinking vs row-wise loops {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`ggplot2`**, **`plotly`** — `install.packages(c("dplyr", "ggplot2", "plotly"))` once.
:::
::: {.quest-learn__row}
**By the end you will**
Add a transformed column with **`mutate()`** and **`log1p()`** (log of 1 + x, safe for zeros) without a loop, and plot depth vs log-count.
:::
::: {.quest-learn__row}
**Apply it later**
Any time AI suggests `for (i in 1:nrow(df))`, ask whether **`mutate()`** across a column is enough — often it is for deterministic transforms on tidy data.
:::
:::

**What you'll learn:** **`mutate()`** adds or changes columns on a data frame. The pipe **`|>`** passes the left side as the first argument to the right side. **`log1p(count)`** is vectorised: R applies it to **every row at once**. **`ggplot()`** + **`aes()`** map variables to axes; **`geom_point()`** draws points.

```{r quest-2}
reef_transects <- reef_transects |> # start with our reprex table
  mutate(log_count = log1p(count)) # new column: log(1 + count), handles zero counts

ggplot(reef_transects, aes(x = depth_m, y = log_count, color = site)) + # map depth, log count, colour by site
  geom_point(size = 2, alpha = 0.85) + # draw one point per row
  theme_minimal(base_size = 13) + # clean theme for slides and web
  labs( # axis and title labels
    title = "Log counts vs depth",
    subtitle = "Vectorised mutate — no explicit loop",
    y = "log1p(count)"
  )
```

```{r quest-2-plotly}
#| echo: false
p2 <- ggplot(reef_transects, aes(
  x = depth_m, y = log_count, color = site,
  text = paste(species, transect_id, sep = " · ")
)) +
  geom_point(size = 2.5, alpha = 0.9) +
  theme_minimal(base_size = 13) +
  labs(y = "log1p(count)", x = "Depth (m)")
plotly::ggplotly(p2, tooltip = "text")
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 3 · ~10 min
:::

### `apply()` — functions on matrix margins {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`tidyr`**, **`tibble`**.
:::
::: {.quest-learn__row}
**By the end you will**
Turn a long species table into a matrix and use **`apply(X, MARGIN, FUN)`** to sum rows (transects) or average columns (species).
:::
::: {.quest-learn__row}
**Apply it later**
Acoustic telemetry matrices, presence–absence grids, and monthly catch tables often start as matrices — **`apply()`** is still what many AI scripts use; you should recognise **`MARGIN = 1`** (rows) vs **`2`** (columns).
:::
:::

**What you'll learn:** **`select()`** keeps columns. **`pivot_wider()`** spreads species counts into columns. **`column_to_rownames()`** moves transect IDs to row names. **`as.matrix()`** makes a numeric matrix. **`apply()`** applies **`sum`** or **`mean`** along margin **`1`** (each row) or **`2`** (each column).

```{r quest-3}
count_matrix <- reef_transects |> # long-format counts
  select(transect_id, species, count) |> # keep keys and response
  pivot_wider(names_from = species, values_from = count) |> # one column per species
  column_to_rownames("transect_id") |> # transect ID becomes row name
  as.matrix() # apply() wants a matrix

transect_totals <- apply(count_matrix, 1, sum) # margin 1: sum each row (one total per transect)
head(transect_totals) # first six transect totals

species_means <- apply(count_matrix, 2, mean) # margin 2: mean each column (one mean per species)
round(species_means, 2) # round for readable output
```

**Check yourself:** Which margin answers *“how many fish per transect?”* Which answers *“average count per species across transects?”*
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 4 · ~10 min
:::

### `split()` and `lapply()` — lists by group {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
`base R` only.
:::
::: {.quest-learn__row}
**By the end you will**
Split counts by site into a list and summarise each element with **`lapply()`**, always getting a **list** back (same length as input).
:::
::: {.quest-learn__row}
**Apply it later**
Per-receiver detection lists, per-animal ID tracks, or per-trip dive logs — ecology is full of “do this for each group.” **`split()` + `lapply()`** is the pattern **`purrr`** improves in Quest 11.
:::
:::

**What you'll learn:** **`split(x, f)`** breaks vector **`x`** into a list by factor **`f`**. **`lapply(list, FUN)`** runs **`FUN`** on each list element. Anonymous functions look like **`function(x) { ... }`** when you have not named the recipe yet.

```{r quest-4}
by_site <- split(reef_transects$count, reef_transects$site) # list of count vectors, one per site

site_summaries <- lapply(by_site, function(x) { # for each site's counts...
  c(mean = mean(x), sd = sd(x), n = length(x)) # return mean, sd, and sample size
}) # result is a list with names North, South, East

site_summaries # inspect each site's summary vector
```

```{r quest-4-viz}
site_mean_df <- tibble::tibble( # small table for plotting
  site = names(site_summaries), # site names from the list
  mean_count = vapply(site_summaries, function(x) x["mean"], numeric(1)) # extract mean from each list element
)

ggplot(site_mean_df, aes(x = reorder(site, mean_count), y = mean_count, fill = site)) + # order bars by height
  geom_col(show.legend = FALSE, width = 0.65) + # bar chart
  scale_fill_manual(values = c(North = "#1f5460", South = "#2a9d8f", East = "#0077b6")) + # reef palette
  theme_minimal(base_size = 13) +
  labs(x = NULL, y = "Mean count", title = "Mean fish count by site")
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 5 · ~10 min
:::

### `sapply()` and the “simplify” trap {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
`base R` only (uses **`by_site`** from Quest 4).
:::
::: {.quest-learn__row}
**By the end you will**
Use **`sapply()`** when every result has the same type and length, and know when AI should **not** use it (unequal lengths → stick with **`lapply()`** or purrr).
:::
::: {.quest-learn__row}
**Apply it later**
When generated code uses **`sapply()`** on heterogeneous model outputs, replace with **`map()`** / **`map_dbl()`** (Quest 12) so type surprises do not break **`bind_rows()`**.
:::
:::

**What you'll learn:** **`sapply()`** is **`lapply()`** plus automatic simplification to a vector or matrix. That is convenient for **`mean`** per site, dangerous when each group returns a different length.

```{r quest-5}
sapply(by_site, mean) # each site returns one number → simplified to named vector

# If summaries had different lengths, sapply() could return an unpredictable shape.
# AI often hides that bug — prefer lapply() or purrr until Quest 11.
```
:::

---

::: {.quest-step}
::: {.quest-step__meta}
Quest 6 · ~10 min · Level 2 — Wrangler
:::

### `across()` — one `summarise()`, many columns {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Compute mean and sd for several numeric columns per **`site`** in one **`summarise()`**, using **`across()`** instead of copy-pasted **`mean(x)`** lines.
:::
::: {.quest-learn__row}
**Apply it later**
Reef monitoring with depth, temperature, and count — **`across(where(is.numeric), ...)`** is the tidy pattern AI should use instead of twelve **`mutate()`** lines.
:::
:::

**What you'll learn:** **`group_by()`** defines groups. **`summarise()`** collapses each group to one row. **`across(cols, .fns)`** applies functions to multiple columns. **`all_of()`** selects columns by name safely. **`.names = "{.col}_{.fn}"`** controls output column names ([Wickham et al., 2023](https://doi.org/10.32614/RJ-2023-023)).

```{r quest-6}
numeric_cols <- c("count", "depth_m") # columns we want summaries for

reef_transects |>
  group_by(site) |> # one row per site in the output
  summarise(
    across( # apply both mean and sd to each numeric column
      all_of(numeric_cols), # only count and depth_m
      list(mean = mean, sd = sd), # named list of functions
      .names = "{.col}_{.fn}" # e.g. count_mean, depth_m_sd
    ),
    .groups = "drop" # ungroup after summarise
  )
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 7 · ~10 min
:::

### Custom functions inside `summarise()` and `across()` {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Write **`cv()`** (coefficient of variation) and use it inside **`summarise()`** to compare count variability by species.
:::
::: {.quest-learn__row}
**Apply it later**
Index of aggregation, CV of CPUE, or detection probability variance — put the ecology in a named function, then let **`across()`** apply it everywhere.
:::
:::

**What you'll learn:** **`cv()`** is **`sd(x) / mean(x)`**. **`arrange(desc(...))`** sorts descending. **`summarise(cv_count = cv(count))`** passes the **`count`** column into **`cv`**.

```{r quest-7}
cv <- function(x) { # coefficient of variation: unitless spread
  sd(x) / mean(x) # sd divided by mean
}

reef_transects |>
  group_by(species) |> # one row per species
  summarise(cv_count = cv(count), .groups = "drop") |> # apply cv to counts in each group
  arrange(desc(cv_count)) # most variable species first
```

**Try:** Group by **`site`**, use **`across(where(is.numeric), cv)`** inside **`summarise()`**.
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 8 · ~10 min
:::

### `group_modify()` — custom data frame per group {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Return the **top two species by total count** within each site using **`group_modify()`**, which passes each group's data as **`.x`**.
:::
::: {.quest-learn__row}
**Apply it later**
“Top receivers per shark,” “best months per tag” — whenever each group needs a **small table** back, not one number, **`group_modify()`** beats a fragile loop.
:::
:::

**What you'll learn:** **`group_modify(.x, ~ { ... })`** runs on each group; **`.x`** is that group's rows. **`slice_max(total, n = 2)`** keeps the two largest **`total`** values ([dplyr slice docs](https://dplyr.tidyverse.org/reference/slice.html)).

```{r quest-8}
top_species_per_site <- reef_transects |>
  group_by(site) |> # split by reef site
  group_modify(~ { # .x is all rows for one site
    .x |>
      group_by(species) |> # species within this site
      summarise(total = sum(count), .groups = "drop") |> # total count per species
      slice_max(total, n = 2) # keep top two species by abundance
  })

top_species_per_site
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 9 · ~10 min
:::

### `.by` — per-group columns without `group_by()` {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`** ≥ 1.1.0.
:::
::: {.quest-learn__row}
**By the end you will**
Add **`mean_count_site`** within each **`site`** while keeping all rows, using **`.by = site`** inside **`mutate()`**.
:::
::: {.quest-learn__row}
**Apply it later**
Compare each transect to the site mean, or each detection to the array mean — **`.by`** keeps long tables long (good before joins or maps).
:::
:::

**What you'll learn:** **`.by = site`** tells **`mutate()`** to compute **`mean(count)`** separately within each site, then broadcast the result to every row in that site. **`select()`** picks columns; **`head()`** shows the first rows.

```{r quest-9}
reef_transects |>
  mutate( # add columns without collapsing rows
    mean_count_site = mean(count), # mean count within each site
    .by = site # grouping column for the calculation above
  ) |>
  select(site, species, count, mean_count_site) |> # keep readable columns
  head(6) # preview
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 10 · ~10 min
:::

### Refactor repeated logic into a function + `group_modify()` {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Encapsulate **`summarise_abundance()`** and apply it to every **`site × species`** combination — the pattern behind many AI-generated “split-apply-combine” scripts.
:::
::: {.quest-learn__row}
**Apply it later**
Give AI **your** function definition in the prompt: *“Use `summarise_abundance()` on each tag ID.”* You stay in control of definitions; AI fills the boilerplate.
:::
:::

**What you'll learn:** **`n_distinct(transect_id)`** counts unique transects. **`group_by(site, species)`** creates cross-classified groups. **`group_modify(~ summarise_abundance(.x))`** applies your function to each group's data frame **`.x`**.

```{r quest-10}
summarise_abundance <- function(df) { # df is one group's rows
  df |>
    summarise( # collapse this group to one row
      total = sum(count), # total fish counted
      n_transects = n_distinct(transect_id), # how many transects contributed
      .groups = "drop"
    )
}

reef_transects |>
  group_by(site, species) |> # every site–species combination
  group_modify(~ summarise_abundance(.x)) # run our recipe on each piece
```
:::

---

::: {.quest-step}
::: {.quest-step__meta}
Quest 11 · ~10 min · Level 3 — Mapper
:::

### `map()` — one output per list element {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`purrr`** — `install.packages("purrr")`.
:::
::: {.quest-learn__row}
**By the end you will**
Split data by site into a list, **`map()`** a mean over each tibble, and read **`~ mean(.x$count)`** (purrr formula shorthand for anonymous functions).
:::
::: {.quest-learn__row}
**Apply it later**
Lists of spatial objects, nested model fits, or per-year data frames — **`map()`** is the idiomatic replacement for **`lapply()`** when you want consistent naming and typed variants ([Henry & Wickham, 2020](https://purrr.tidyverse.org/)).
:::
:::

**What you'll learn:** **`group_split()`** returns a list of data frames. **`group_keys()`** + **`pull(site)`** gives site names for the list. **`map(list, ~ ...)`** applies the expression to each element; **`.x`** is the current element. **`~`** is shorthand for **`function(.x) ...`**.

```{r quest-11}
site_lists <- reef_transects |>
  group_by(site) |> # prepare grouping
  group_split() # list of three data frames

names(site_lists) <- reef_transects |>
  group_by(site) |>
  group_keys() |> # one row per site
  pull(site) # character vector of site names

means_list <- map(site_lists, ~ mean(.x$count)) # mean count per site's tibble

means_list
```

```{r quest-11-interactive}
#| echo: false
map_df <- tibble::tibble(
  site = names(means_list),
  mean_count = unlist(means_list)
)
plot_11 <- ggplot(map_df, aes(x = site, y = mean_count, fill = site)) +
  geom_col(show.legend = FALSE) +
  theme_minimal() +
  labs(title = "map() output — one mean per site list element")
plotly::ggplotly(plot_11)
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 12 · ~10 min
:::

### Typed maps: `map_dbl()`, `map_chr()` {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`purrr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Return a **double vector** and a **character vector** directly — no **`sapply()`** simplify surprises.
:::
::: {.quest-learn__row}
**Apply it later**
When binding model AICs or tag IDs, insist AI use **`map_dbl()`** / **`map_chr()`** so **`c()`** and **`bind_rows()`** get predictable types.
:::
:::

**What you'll learn:** **`map_dbl()`** always returns a numeric vector of length **`length(list)`**. **`map_chr()`** returns characters. **`paste(..., collapse = ", ")`** joins strings.

```{r quest-12}
map_dbl(site_lists, ~ mean(.x$count)) # numeric vector of means

map_chr(site_lists, ~ paste(unique(.x$transect_id), collapse = ", ")) # transect IDs as one string per site
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 13 · ~10 min
:::

### `map2()` — two parallel vectors {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`purrr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Multiply each species' total count by a species-specific mean weight using **`map2_dbl()`**, with **`.x`** and **`.y`** as the paired elements.
:::
::: {.quest-learn__row}
**Apply it later**
Length–weight regressions applied per species, matching survey IDs to metadata rows, or pairing lat with lon lists — if AI uses **`map()`** where two vectors align, ask for **`map2()`**.
:::
:::

**What you'll learn:** **`map2_dbl(.x, .y, ~ ...)`** walks two vectors in parallel. Inside **`~`**, **`.x`** is the species name, **`.y`** is the total count. **`mean_weight_kg[[.x]]`** looks up weight by name.

```{r quest-13}
species_tbl <- reef_transects |>
  group_by(species) |>
  summarise(total = sum(count), mean_depth = mean(depth_m), .groups = "drop") # one row per species

mean_weight_kg <- c( # named vector: kg per individual by species
  Parrotfish = 0.35, Surgeonfish = 0.55, Grouper = 2.1,
  Snapper = 0.8, Damselfish = 0.05
)

species_tbl |>
  mutate(
    biomass_kg = map2_dbl( # parallel walk over two columns
      species, total, # .x = species name, .y = total count
      ~ .y * mean_weight_kg[[.x]] # count × species-specific weight
    )
  )
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 14 · ~10 min
:::

### `pmap()` — one row at a time {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`purrr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Classify each transect's depth as shallow / mid / deep with **`pmap_chr()`**, the row-wise mapper.
:::
::: {.quest-learn__row}
**Apply it later**
Complex **`ifelse()`** ladders on many columns — **`pmap()`** is clearer and easier for you (and AI) to test row by row.
:::
:::

**What you'll learn:** **`distinct()`** keeps unique rows. **`pmap_chr(list(col), ~ ...)`** passes each row's values as **`..1`**, **`..2`**, …; with one column, **`..1`** or **`.x`** in a one-argument function is the depth value.

```{r quest-14}
reef_transects |>
  distinct(transect_id, site, depth_m) |> # one row per transect
  mutate(
    depth_class = pmap_chr( # character result, one label per row
      list(depth_m), # columns to pass row-wise (here, just depth)
      ~ if (.x < 6) "shallow" else if (.x < 11) "mid" else "deep" # .x is depth for this row
    )
  )
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 15 · ~10 min
:::

### `map_dfr()` — stack many data frames {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`purrr`**, **`ggplot2`**, **`tibble`**.
:::
::: {.quest-learn__row}
**By the end you will**
Simulate eight transects by mapping **`simulate_transect()`** over IDs and row-binding with **`map_dfr()`**.
:::
::: {.quest-learn__row}
**Apply it later**
Monte Carlo abundance, bootstrap transects, or “what if we sampled again?” — **`map_dfr()`** is the standard simulation stacker.
:::
:::

**What you'll learn:** **`rpois(n, lambda)`** draws Poisson counts. **`simulate_transect()`** returns a small tibble. **`map_dfr(ids, fun, ...)`** calls **`fun`** for each ID and **`bind_rows()`** the results. **`sprintf()`** formats ID strings.

```{r quest-15}
simulate_transect <- function(transect_id, lambda = 8) { # one simulated transect
  tibble::tibble(
    transect_id = transect_id, # label this draw
    count = rpois(5, lambda = lambda) # five species counts at this lambda
  )
}

sim_ids <- sprintf("sim_%02d", 1:8) # eight simulation IDs

sim_data <- map_dfr(sim_ids, simulate_transect, lambda = 10) # stack eight tibbles

ggplot(sim_data, aes(x = transect_id, y = count)) + # plot simulated draws
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Eight simulated transects via map_dfr()")
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 16 · ~10 min
:::

### `safely()` — errors become data {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`purrr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Wrap a fragile ratio in **`safely()`** so failed divisions return **`error`** instead of stopping your script — then inspect with **`map()`**.
:::
::: {.quest-learn__row}
**Apply it later**
Batch-reading receiver folders, fitting GLMMs per tag, parsing messy CSVs — **`safely()`** + **`map()`** is how you QC 200 files without **`tryCatch()`** soup. Tell AI: *“wrap the reader in `safely()`.”*
:::
:::

**What you'll learn:** **`safely(f)`** returns a new function that always gives a list with **`result`** and **`error`**. **`map_lgl()`** maps to logical. **`is.null(.x$error)`** means success.

```{r quest-16}
safe_ratio <- safely(function(num, den) { # safely wraps our division
  num / den # fails when den is 0
})

tests <- list( # three test cases
  safe_ratio(10, 2), # ok
  safe_ratio(3, 0), # error captured
  safe_ratio(NA, 5) # may return NA in result
)

map_lgl(tests, ~ is.null(.x$error)) # TRUE where no error

map(tests, ~ .x$result) # extract results (NULL where error)
```
:::

---

::: {.quest-step}
::: {.quest-step__meta}
Quest 17 · ~10 min · Level 4 — Parallel diver
:::

### `furrr` — parallel `map()` when tasks are independent {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`purrr`**, **`future`**, **`furrr`** — `install.packages(c("future", "furrr"))`.
:::
::: {.quest-learn__row}
**By the end you will**
Fit a Poisson **`glm()`** per species on separate workers with **`future_map()`**, after **`plan(multisession)`**.
:::
::: {.quest-learn__row}
**Apply it later**
Bootstrap CIs, many MCMC restarts, or per-AnimalID home-range models — parallelise only when each task is **independent** ([Bengtsson, 2021](https://doi.org/10.32614/RJ-2021-048)). Ask AI for **`furrr`**, not random **`parallel::mclapply`** on Windows.
:::
:::

**What you'll learn:** **`plan(multisession, workers = 2)`** starts a small worker pool. **`future_map()`** like **`map()`** but parallel. **`tryCatch()`** catches errors inside **`fit_poisson`**. **`glm(..., family = poisson)`** fits count vs depth.

```{r quest-17}
if (requireNamespace("furrr", quietly = TRUE)) { # only run if furrr is installed
  library(furrr) # parallel purrr
  plan(multisession, workers = 2) # two background R sessions

  species_splits <- reef_transects |>
    group_split(species) # one tibble per species

  fit_poisson <- function(df) { # one glm per species table
    tryCatch(
      glm(count ~ depth_m, data = df, family = poisson), # Poisson regression
      error = function(e) NULL # return NULL instead of stopping
    )
  }

  models <- future_map(species_splits, fit_poisson) # fit models in parallel

  map_lgl(models, ~ !is.null(.x)) # which species fit successfully?
} else {
  message("Install furrr to run Quest 17 locally: install.packages('furrr')")
}
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 18 · ~10 min
:::

### `walk()` — map for side effects only {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`dplyr`**, **`purrr`**.
:::
::: {.quest-learn__row}
**By the end you will**
Print a one-line site report for each list element with **`walk()`**, which discards return values (unlike **`map()`**).
:::
::: {.quest-learn__row}
**Apply it later**
**`ggsave()`** per tag, **`write_csv()`** per trip, **`message()`** QC logs — **`walk2()`** pairs IDs with plots. AI loves **`map()`** for this; **`walk()`** is the correct tool.
:::
:::

**What you'll learn:** **`walk(list, ~ ...)`** runs code for each element and returns **`list`** invisibly. **`message()`** prints to the console without stopping. **`paste0()`** builds strings. **`n_distinct()`** counts unique values.

```{r quest-18}
report_site <- function(df) { # one-line summary string
  paste0(
    unique(df$site)[1], # site name
    ": ",
    n_distinct(df$transect_id), # how many transects
    " transects, total count = ",
    sum(df$count) # sum of counts
  )
}

walk(site_lists, ~ message(report_site(.x))) # message each report; return value ignored

invisible(NULL) # chunk returns quietly in HTML
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 19 · ~10 min
:::

### data.table: `[i, j, by]` {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`data.table`** — `install.packages("data.table")`.
:::
::: {.quest-learn__row}
**By the end you will**
Compute mean count and **`.N`** (row count) per **`site`** in one **`DT[i, j, by]`** expression.
:::
::: {.quest-learn__row}
**Apply it later**
Million-row acoustic detection tables — data.table is what you want when AI's dplyr pipeline is correct but slow. Learn to **read** `[ , .(...), by = ...]` even if you write dplyr day-to-day ([Dowle & Srinivasan, 2024](https://doi.org/10.32614/RJ-2024-001)).
:::
:::

**What you'll learn:** **`as.data.table()`** converts without copying when possible. Syntax **`DT[i, j, by]`**: **`i`** filters rows (empty = all), **`j`** computes columns, **`by`** groups. **`.(...)`** lists output columns. **`.N`** is group size. **`order(-mean_count)`** sorts descending.

```{r quest-19}
dt <- as.data.table(reef_transects) # data.table version of our reprex

dt[
  , # i: no row filter — use all rows
  .( # j: columns to return
    mean_count = mean(count), # average count in group
    n = .N # number of rows in group
  ),
  by = site # group by site
][order(-mean_count)] # sort result by mean_count descending
```
:::

::: {.quest-step}
::: {.quest-step__meta}
Quest 20 · ~10 min
:::

### `.SD` — group subsets in data.table {.unnumbered}

<span class="xp-pill">+50 XP</span>

::: {.quest-learn}
::: {.quest-learn__row}
**Packages**
**`data.table`**, **`ggplot2`**, **`plotly`**, **`tibble`** (for **`as_tibble()`** before ggplot).
:::
::: {.quest-learn__row}
**By the end you will**
Summarise **`mean(count)`** and **`mean(depth_m)`** by **`site`** and **`species`**, then plot grouped bars — same mental model as **`group_by()` + `summarise()`**.
:::
::: {.quest-learn__row}
**Apply it later**
When AI pastes data.table code into your telemetry pipeline, recognise **`.(site, species)`** as grouping and **`.SD`** as “columns available in this group” — then decide if you keep it or translate to dplyr you can maintain.
:::
:::

**What you'll learn:** **`by = .(site, species)`** groups on two columns. **`mean(count)`** in **`j`** is evaluated within each group. **`as_tibble()`** converts for **`ggplot()`**. **`geom_col(position = "dodge")`** places bars side by side.

```{r quest-20}
dt[
  ,
  .(
    mean_count = mean(count), # average abundance
    mean_depth = mean(depth_m) # average depth
  ),
  by = .(site, species) # cross of site and species
]

plot_20 <- dt[
  , .(mean_count = mean(count)), # simpler summary for plotting
  by = .(site, species)
] |>
  as_tibble() |> # ggplot prefers tibble
  ggplot(aes(x = species, y = mean_count, fill = site)) + # dodged bars by site
  geom_col(position = "dodge") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1)) + # readable species names
  labs(title = "Group means — data.table → ggplot")

plotly::ggplotly(plot_20)
```
:::

---

## Season 1 complete — you beat the paste buffer

Twenty quests ≈ **1,000 XP**. You can now **read** functional ecology code, **critique** AI output, and **request** the right tool (`map2`, `safely`, `walk`, `[ , , by]`).

### Roadmap · quests 21–100

| Season | Quests | Focus | ~time |
|--------|--------|--------|-------|
| 2 | 21–30 | `map_dfc`, list-columns, **`unnest()`** | 1 week |
| 3 | 31–40 | **`reduce()`**, spatial list columns | 1 week |
| 4 | 41–50 | **`furrr`** bootstrap CIs per reef | 1 week |
| 5 | 51–60 | data.table joins on tag metadata | 1 week |
| 6 | 61–70 | **`possibly()`**, **`quietly()`**, failed models | 1 week |
| 7 | 71–80 | purrr + **sf** per-animal ranges | 2 weeks |
| 8 | 81–90 | temporal **`slide()`** patterns | 2 weeks |
| 9 | 91–95 | your first package function | 1 week |
| 10 | 96–100 | capstone: refactor telemetry functionally | 1 week |

::: {.callout-important appearance="simple"}
## Reprex for humans and AI

When you ask for help — from a forum or from a model — paste **data + sessionInfo()**, not a screenshot:

``` r
set.seed(427)
# ... reef_transects chunk from above ...
sessionInfo()
```

Prompt tip: *“Use my `summarise_abundance()` function; do not introduce new column names.”*
:::

## References

- Bengtsson, H. (2021). A unifying framework for parallel and distributed processing in R using futures. *The R Journal*, 13(2). <https://doi.org/10.32614/RJ-2021-048>
- Bransford, J. D., Brown, A. L., & Cocking, R. R. (2000). *How people learn*. National Academies Press.
- Dowle, M., & Srinivasan, A. (2024). data.table: Extension of `data.frame`. *The R Journal*. <https://doi.org/10.32614/RJ-2024-001>
- Henry, L., & Wickham, H. (2020). *purrr*: Functional programming tools. <https://purrr.tidyverse.org/>
- Mayer, R. E. (2009). Multimedia learning (2nd ed.). Cambridge University Press.
- Roediger, H. L., III, & Butler, A. C. (2011). The critical role of retrieval practice in long-term retention. *Trends in Cognitive Sciences*, 15(1), 20–27. <https://doi.org/10.1016/j.jics.2010.09.004>
- Rosenshine, B. (2012). Principles of instruction: Research-based strategies that all teachers should know. *American Educator*, 36(1), 12–19. <https://doi.org/10.1177/0031721712452808>
- Wickham, H. (2014). *Advanced R* — Functional programming. <https://adv-r.hadley.nz/functional-programming.html>
- Wickham, H., François, R., Henry, L., Müller, K., & Vaughan, D. (2023). dplyr: A grammar of data manipulation. *The R Journal*, 15(3). <https://doi.org/10.32614/RJ-2023-023>
- The Big Book of R — Functional programming. <https://bigbookofr.com/functional-programming.html>
