---
title: "Functional programming · Advanced (data.table)"
description: "Advanced · data.table"
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Advanced (data.table)
`[i, j, by]` when data get large.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
DTreef <- as.data.table(reef)
DTfish <- as.data.table(fishery)
DTac <- as.data.table(acoustic)
DTbio <- as.data.table(biolog)
```

## Q1 · as.data.table() mindset {#q01}
```{r q01}
print(DTreef) # data.table print method
class(DTreef) # data.table and data.frame
dim(DTreef) # rows and columns
```

## Q2 · [i, j, by] core {#q02}
```{r q02}
DTreef[site == "North", .(mean_count = mean(count)), by = species]
```

**Ex 2.**

```{r q02-fishery}
DTfish[port == "Cairns", .(mean_t = mean(tonnes)), by = species]
```

**Ex 2.**

```{r q02-acoustic}
DTac[reef_zone == "Lagoon", .(total = sum(detections)), by = receiver]
```

**Ex 2.**

```{r q02-biolog}
DTbio[, .(max_depth = max(depth_m)), by = burst_id][1:5]
```

## Q3 · .N and .I {#q03}
```{r q03}
DTac[, .N, by = animal_id] # rows per tag
DTac[, .(max_det = max(detections)), by = animal_id] # peak detection per tag
```

## Q4 · .SD group columns {#q04}
```{r q04}
DTbio[, lapply(.SD, mean), by = burst_id, .SDcols = c("depth_m", "pitch_deg")]
```

**Ex 2.**

```{r q04-reef}
DTreef[, lapply(.SD, mean), by = site, .SDcols = c("count", "depth_m")]
```

**Ex 2.**

```{r q04-fishery}
DTfish[, .(min_t = min(tonnes), max_t = max(tonnes)), by = port]
```

## Q5 · := assign by reference {#q05}
```{r q05}
DTfish[, kg := tonnes * 1000] # new column in place
head(DTfish[, .(species, tonnes, kg)])
```

## Q6 · setorder() fast sort {#q06}
```{r q06}
setorder(DTbio, burst_id, seconds) # in-place sort
head(DTbio, 5)
```

## Q7 · Keys and binary search {#q07}
```{r q07}
DTac_k <- copy(DTac) # avoid mutating shared object
setkey(DTac_k, animal_id)
DTac_k["A101"]
```

## Q8 · merge() joins {#q08}
```{r q08}
ports <- data.table(port = c("Cairns", "Broome", "Mooloolaba"),
                      region = c("GBR", "NW", "SEQ"))
merge(DTfish, ports, by = "port", all.x = TRUE) # keep all landings
```

**Ex 2.**

```{r q08-acoustic}
meta <- data.table(animal_id = c("A101", "A102", "A103"), cohort = c(2023, 2023, 2024))
merge(DTac, meta, by = "animal_id", all.x = TRUE)[1:6]
```

**Ex 2.**

```{r q08-reef}
site_meta <- data.table(site = c("North", "South", "East"), mgmt = c("MPA", "Open", "MPA"))
merge(DTreef, site_meta, by = "site")[1:6]
```

## Q9 · rbindlist() stack {#q09}
```{r q09}
north <- DTreef[site == "North"] # subset table
south <- DTreef[site == "South"]
rbindlist(list(north, south), use.names = TRUE) # combined table
```

## Q10 · dcast() wide {#q10}
```{r q10}
dcast(DTac, animal_id ~ receiver, value.var = "detections", fun.aggregate = sum, fill = 0)
```

## Q11 · melt() long {#q11}
```{r q11}
wide <- dcast(DTfish[year == 2020], year ~ species, value.var = "tonnes", fill = 0)
melt(wide, id.vars = "year", variable.name = "species", value.name = "tonnes")
```

## Q12 · shift() like lag {#q12}
```{r q12}
DTbio[, depth_change := depth_m - shift(depth_m), by = burst_id]
DTbio[!is.na(depth_change), .(burst_id, seconds, depth_change)][1:8]
```

**Ex 2.**

```{r q12-fishery}
DTf <- copy(DTfish) # do not mutate capstone copy
DTf[, year := as.integer(year)] # ensure sort order
setorder(DTf, species, year)
DTf[, tonnes_lag := shift(tonnes), by = species]
DTf[!is.na(tonnes_lag), .(species, year, tonnes, tonnes_lag)][1:6]
```

**Ex 2.**

```{r q12-acoustic}
DTac[, det_lag := shift(detections), by = animal_id]
DTac[!is.na(det_lag), .(animal_id, receiver, detections, det_lag)][1:6]
```

## Q13 · fcase() like case_when {#q13}
```{r q13}
DTfish[, size_bin := fcase(
  tonnes < 30, "small",
  tonnes < 70, "medium",
  default = "large"
)]
DTfish[, .N, by = size_bin] # bin counts
```

**Ex 2.**

```{r q13-reef}
DTreef[, depth_bin := fcase(depth_m < 5, "shallow", depth_m < 12, "mid", default = "deep")]
DTreef[, .N, by = depth_bin]
```

**Ex 2.**

```{r q13-acoustic}
DTac[, activity := fcase(detections == 0, "none", detections < 4, "low", default = "high")]
DTac[, .N, by = activity]
```

## Q14 · Grouped chain ; {#q14}
```{r q14}
east <- DTreef[site == "East"]
{
  east[, .(mean_depth = mean(depth_m), total_fish = sum(count))];
  east[, .(species, count)][order(-count)]
}
```

## Q15 · fread() big CSV {#q15}
```{r q15}
reef_fread <- fread(data_path("reef_transects.csv"))
nrow(reef_fread) # should match DTreef
```

## Q16 · Telemetry aggregation {#q16}
```{r q16}
tel <- DTac[, .(total = sum(detections), n = .N), by = .(reef_zone, receiver)]
tel[order(-total)][1:10] # top ten cells
```

## Q17 · Fishery by year {#q17}
```{r q17}
DTfish[, .(tonnes = sum(tonnes), records = .N), by = year][order(year)]
```

## Q18 · Biologging rolling mean {#q18}
```{r q18}
DTbio[, roll_depth := (shift(depth_m, 1, type = "lag") + depth_m + shift(depth_m, 1, type = "lead")) / 3, by = burst_id]
DTbio[!is.na(roll_depth), .(burst_id, seconds, depth_m, roll_depth)][1:8]
```

## Q19 · .BY in j {#q19}
```{r q19}
DTac[, .(
  zone = .BY[[1]],
  total = sum(detections)
), by = reef_zone]
```

## Q20 · data.table capstone {#q20}
```{r q20}
meta <- data.table(site = c("North", "South", "East"), mgmt = c("MPA", "Open", "MPA"))
reef_sum <- DTreef[, .(total = sum(count), species_n = uniqueN(species)), by = site]
out <- merge(reef_sum, meta, by = "site")
ggplot(as.data.frame(out), aes(x = site, y = total, fill = mgmt)) +
  geom_col() +
  labs(title = "Reef fish totals by site and management", y = "Sum of counts")
```

{{< include ../_includes/learn-more-fp.md >}}
