---
title: "Functional programming · Advanced (data.table)"
description: "Advanced track — twenty quests on [i,j,by], joins, shift, and fast aggregation."
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Advanced (data.table)

**data.table** treats columns as memory you reference, not copy — syntax **`[i, j, by]`** expresses filter, compute, and group in one line. Experienced analysts often pick one dialect per workflow (**dplyr** *or* **data.table**) for readability.

Keys, binary search, and **`:=`** assignment by reference are performance tools; the deeper win is explicit grouping via **`by`** and **`.SD`**. Predict row counts per group before running code — the same hypothesis-check habit from scientific reasoning ([Klahr & Dunbar, 1988](https://doi.org/10.1016/0166-4115(88)90004-2)).

When CSVs grow large, **`fread()`** replaces **`read.csv()`**; we note it in a quest but keep setup on **`read.csv()`** so renders stay portable.

Core syntax quests repeat on **reef, fishery, acoustic, and biologging** tables so you read the expression, not one memorised column name.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
DTreef <- as.data.table(reef)
DTfish <- as.data.table(fishery)
DTac <- as.data.table(acoustic)
DTbio <- as.data.table(biolog)
```

## Q1 · as.data.table() mindset {#q01}

**`as.data.table()`** converts data frames without dropping columns; printing shows head and tail of large tables.

data.table is still tabular data — but `[` overloads filtering, grouping, and assignment.

Convert reef data and inspect structure.

```{r q01}
print(DTreef)                                     # data.table print method
class(DTreef)                                           # data.table and data.frame
dim(DTreef)                                             # rows and columns
```

## Q2 · [i, j, by] core {#q02}

The template **`DT[i, j, by]`** uses **i** to filter rows, **j** to compute, **by** to group. The `.()` helper builds j expressions with names.

Read expressions aloud before running — catches AI mistakes like filtering on wrong column.

Mean count by species for North site transects only.

```{r q02}
DTreef[site == "North", .(mean_count = mean(count)), by = species]
```

**Example 2 — fishery:** mean **`tonnes`** by **`species`** for **`port == "Cairns"`**.

```{r q02-fishery}
DTfish[port == "Cairns", .(mean_t = mean(tonnes)), by = species]
```

**Example 3 — acoustic:** sum **`detections`** by **`receiver`** in **`reef_zone == "Lagoon"`**.

```{r q02-acoustic}
DTac[reef_zone == "Lagoon", .(total = sum(detections)), by = receiver]
```

**Example 4 — biologging:** max **`depth_m`** by **`burst_id`**.

```{r q02-biolog}
DTbio[, .(max_depth = max(depth_m)), by = burst_id][1:5]
```

## Q3 · .N and .I {#q03}

**.N** counts rows in the current group; **.I** holds indices for within-group positions.

These symbols replace verbose split-apply counting in base R.

Count acoustic rows per animal and pull each tag's maximum detection.

```{r q03}
DTac[, .N, by = animal_id]                        # rows per tag
DTac[, .(max_det = max(detections)), by = animal_id]    # peak detection per tag
```

## Q4 · .SD group columns {#q04}

**.SD** is the Subset of Data for each group — all non-group columns inside j.

Combine with `.SDcols` to apply functions across selected numeric columns.

Mean depth and pitch per biologging burst.

```{r q04}
DTbio[, lapply(.SD, mean), by = burst_id, .SDcols = c("depth_m", "pitch_deg")]
```

**Example 2 — reef:** mean **`count`** and **`depth_m`** per **`site`** via **`.SD`**.

```{r q04-reef}
DTreef[, lapply(.SD, mean), by = site, .SDcols = c("count", "depth_m")]
```

**Example 3 — fishery:** min and max **`tonnes`** per **`port`**.

```{r q04-fishery}
DTfish[, .(min_t = min(tonnes), max_t = max(tonnes)), by = port]
```

## Q5 · := assign by reference {#q05}

**`:=`** adds or updates columns in place without copying the whole table — fast but side-effectful.

Start with simple global assignments before grouped := patterns.

Add kilograms column to fishery landings.

```{r q05}
DTfish[, kg := tonnes * 1000]                     # new column in place
head(DTfish[, .(species, tonnes, kg)])                  # verify conversion
```

## Q6 · setorder() fast sort {#q06}

**`setorder()`** sorts by reference; sorting biologging by burst and second prepares time-series ops.

Unlike dplyr arrange, setorder mutates the table — copy first if you need the original order.

Order biolog by burst_id then seconds.

```{r q06}
setorder(DTbio, burst_id, seconds)                # in-place sort
head(DTbio, 5)                                          # inspect first rows
```

## Q7 · Keys and binary search {#q07}

A **key** indexes sorted column(s) — subsetting `DT[id]` uses binary search without merge syntax.

Set keys on animal_id for acoustic lookups.

Key acoustic table and extract animal A101.

```{r q07}
DTac_k <- copy(DTac)                              # avoid mutating shared object
setkey(DTac_k, animal_id)                               # sort and index
DTac_k["A101"]                                          # fast row subset
```

## Q8 · merge() joins {#q08}

**`merge()`** joins tables; with keys set, joins can be faster but always check nrow after.

Attach port region metadata to fishery landings.

```{r q08}
ports <- data.table(port = c("Cairns", "Broome", "Mooloolaba"),
                      region = c("GBR", "NW", "SEQ"))
merge(DTfish, ports, by = "port", all.x = TRUE)         # keep all landings
```

**Example 2 — acoustic:** join **`animal_id`** to tag **`cohort`** metadata.

```{r q08-acoustic}
meta <- data.table(animal_id = c("A101", "A102", "A103"), cohort = c(2023, 2023, 2024))
merge(DTac, meta, by = "animal_id", all.x = TRUE)[1:6]
```

**Example 3 — reef:** join **`site`** management onto transects (many rows per site).

```{r q08-reef}
site_meta <- data.table(site = c("North", "South", "East"), mgmt = c("MPA", "Open", "MPA"))
merge(DTreef, site_meta, by = "site")[1:6]
```

## Q9 · rbindlist() stack {#q09}

**`rbindlist()`** binds a list of data.tables quickly — stack survey legs or batches.

Use `fill = TRUE` when column sets differ slightly.

Stack North and South reef subsets.

```{r q09}
north <- DTreef[site == "North"]                  # subset table
south <- DTreef[site == "South"]                        # another subset
rbindlist(list(north, south), use.names = TRUE)         # combined table
```

## Q10 · dcast() wide {#q10}

**`dcast()`** pivots long to wide — detection totals by animal and receiver.

Provide fun.aggregate when duplicate keys exist.

Wide matrix of summed detections.

```{r q10}
dcast(DTac, animal_id ~ receiver, value.var = "detections", fun.aggregate = sum, fill = 0)
```

## Q11 · melt() long {#q11}

**`melt()`** unpivots wide tables — inverse of dcast for export pipelines.

Set id.vars to columns that identify rows.

Melt wide 2020 landings by species columns.

```{r q11}
wide <- dcast(DTfish[year == 2020], year ~ species, value.var = "tonnes", fill = 0)
melt(wide, id.vars = "year", variable.name = "species", value.name = "tonnes")
```

## Q12 · shift() like lag {#q12}

**`shift()`** moves values within groups — lag and lead for time series inside bursts.

Negative n shifts forward (lead); default n is one step lag.

Depth change between consecutive seconds per burst.

```{r q12}
DTbio[, depth_change := depth_m - shift(depth_m), by = burst_id]
DTbio[!is.na(depth_change), .(burst_id, seconds, depth_change)][1:8]
```

**Example 2 — fishery:** **`shift(tonnes)`** within **`species`** after **`setorder(species, year)`** (consecutive records).

```{r q12-fishery}
DTf <- copy(DTfish)                                   # do not mutate capstone copy
DTf[, year := as.integer(year)]                         # ensure sort order
setorder(DTf, species, year)
DTf[, tonnes_lag := shift(tonnes), by = species]
DTf[!is.na(tonnes_lag), .(species, year, tonnes, tonnes_lag)][1:6]
```

**Example 3 — acoustic:** **`shift(detections)`** within **`animal_id`** (ordered as in file).

```{r q12-acoustic}
DTac[, det_lag := shift(detections), by = animal_id]
DTac[!is.na(det_lag), .(animal_id, receiver, detections, det_lag)][1:6]
```

## Q13 · fcase() like case_when {#q13}

**`fcase()`** evaluates vectorised conditions top-down — fast alternative to dplyr case_when inside j.

Use default= for a catch-all bin.

Classify fishery landings into size bins.

```{r q13}
DTfish[, size_bin := fcase(
  tonnes < 30, "small",
  tonnes < 70, "medium",
  default = "large"
)]
DTfish[, .N, by = size_bin]                             # bin counts
```

**Example 2 — reef:** depth class with **`fcase`**.

```{r q13-reef}
DTreef[, depth_bin := fcase(depth_m < 5, "shallow", depth_m < 12, "mid", default = "deep")]
DTreef[, .N, by = depth_bin]
```

**Example 3 — acoustic:** activity flag from **`detections`**.

```{r q13-acoustic}
DTac[, activity := fcase(detections == 0, "none", detections < 4, "low", default = "high")]
DTac[, .N, by = activity]
```

## Q14 · Grouped chain ; {#q14}

Curly braces with semicolons run multiple j expressions on the same filtered table — compact EDA.

Keep chains short and commented when teaching.

East reef: overall means then top species counts.

```{r q14}
east <- DTreef[site == "East"]
{
  east[, .(mean_depth = mean(depth_m), total_fish = sum(count))];
  east[, .(species, count)][order(-count)]
}
```

## Q15 · fread() big CSV {#q15}

**`fread()`** reads large CSVs faster than read.csv from the same project-relative path.

If fread is unavailable, read.csv plus as.data.table still works.

Demonstrate fread on reef path with read.csv fallback.

```{r q15}
reef_fread <- fread(data_path("reef_transects.csv"))    # fast read; same path as setup
nrow(reef_fread)                                        # should match DTreef
```

## Q16 · Telemetry aggregation {#q16}

Group acoustic detections by reef_zone and receiver — standard passive telemetry summaries.

Use .() in j for named columns ready for plotting.

Sum detections and count rows per zone-receiver cell.

```{r q16}
tel <- DTac[, .(total = sum(detections), n = .N), by = .(reef_zone, receiver)]
tel[order(-total)][1:10]                                # top ten cells
```

## Q17 · Fishery by year {#q17}

Annual summaries with by = year mirror agency reports — sum tonnes and count records separately.

Records and biomass answer different scientific questions.

Tonnes and landing record counts per year.

```{r q17}
DTfish[, .(tonnes = sum(tonnes), records = .N), by = year][order(year)]
```

## Q18 · Biologging rolling mean {#q18}

Rolling means smooth depth noise along bursts — three-point centred window using shift sums.

Production code may use frollmean; here we teach the lag-lead pattern explicitly.

Centred three-second mean depth per burst.

```{r q18}
DTbio[, roll_depth := (shift(depth_m, 1, type = "lag") + depth_m + shift(depth_m, 1, type = "lead")) / 3, by = burst_id]
DTbio[!is.na(roll_depth), .(burst_id, seconds, depth_m, roll_depth)][1:8]
```

## Q19 · .BY in j {#q19}

**.BY** exposes current by-group values inside j — handy for labels without repeating columns.

.BY is a list; index the first element for single-column groups.

Summarise detections with explicit zone label from .BY.

```{r q19}
DTac[, .(
  zone = .BY[[1]],
  total = sum(detections)
), by = reef_zone]
```

## Q20 · data.table capstone {#q20}

Capstone combines aggregation, merge, and ggplot on reef totals — meeting-ready output you verify before AI optimises.

Check nrow after merge and that grouping columns match field definitions.

Site totals joined to management metadata and plotted.

```{r q20}
meta <- data.table(site = c("North", "South", "East"), mgmt = c("MPA", "Open", "MPA"))
reef_sum <- DTreef[, .(total = sum(count), species_n = uniqueN(species)), by = site]
out <- merge(reef_sum, meta, by = "site")
ggplot(as.data.frame(out), aes(x = site, y = total, fill = mgmt)) +
  geom_col() +
  labs(title = "Reef fish totals by site and management", y = "Sum of counts")
```

