---
title: "Residency and movement for acoustic telemetry"
date: 2026-06-21
categories: [movement, acoustic-telemetry]
description: "Defining residency and movement from acoustic detections using dplyr."
---

21-06-2026 10:44pm

Hiya folks! My website may be written by AI, but this is not, so enjoy.

It's me, your internet friend, Pablo :). If you're reading this, thanks for exploring my first R tutorial!

I thought I'd kick it off with something easy(ish), yet useful - movement for acoustic telemetry.

I actually only learned about the data type in my third year of studying animal ecology, and was really interested in the many ways you could utilise the data, as well as the fact tags can last up to 10 years! It helps being an Aussie, we have IMOS - the Integrated Marine Observing System, which holds one of the largest repositories of animal movement in the world.

One of the many devious tasks associated with wrestling acoustic telemetry, is categorising movement. Relying on packages can simplify life, until they break....

So, I wrote some simple code in dplyr to define non-residency (movement between two detections), and residency (multiple detections at one receiver / array).

Let's begin by downloading an IMOS toolkit REMORA (https://github.com/IMOS-AnimalTracking/remora), and loading some dummy data:

```{r}
# requires a semi-newest version of R, and your ability to install this
# install.packages("remotes")
# install.packages('pacman')
# remotes::install_github('IMOS-AnimalTracking/remora',
#                         build_vignettes = TRUE, # vignettes offline
#                         dependencies = TRUE) # dependecies offline

library(remora)
library(tidyverse)
library(terra)
library(ggspatial)
library(scico)

## Example dataset that has undergone quality control using the `runQC()` function
data("TownsvilleReefQC")

## Only retain detections flagged as 'valid' and 'likely valid' (Detection_QC 1 and 2)
dat <- 
  TownsvilleReefQC %>% 
  tidyr::unnest(cols = QC) %>%
  dplyr::ungroup() %>% 
  filter(Detection_QC %in% c(1,2))


# ggplot it baby
dat1 <- dat %>% 
  group_by(transmitter_id, station_name, installation_name, receiver_deployment_longitude, receiver_deployment_latitude) %>% 
  summarise(num_det = n())  %>% 
  ungroup

  ggplot(dat1) +
  annotation_map_tile('cartolight') +
  geom_spatial_point(aes(x = receiver_deployment_longitude, y = receiver_deployment_latitude, 
                         size = num_det, colour = installation_name), crs = 4326) +
  facet_wrap(~transmitter_id, ncol = 2) +
  labs(x = "Longitude", y = "Latitude", color = "Installation Name" , size = "Number of\nDetections") +
  theme_bw() +
  scale_colour_scico_d(palette = 'lajolla')

# we can also look at detections over time
  dat %>% 
  mutate(date = as.Date(detection_datetime)) %>% 
  group_by(transmitter_id, date, installation_name) %>% 
  summarise(num_det = n()) %>% 
  ggplot(aes(x = date, y = transmitter_id, color = installation_name, size = num_det)) +
  labs(x = "Date", y = NULL, color = "Installation Name", size = "Number of\nDetections") +
  geom_point() +
  theme_bw()

```

Great, we have around 3 years of data across a few islands on the great barrier reef for 5 tags. 
Let's get to making non-residency, aka movement.

Movement uses two detections, and denotes when it leaves one (departure), and arrives at another (arrival). From this information, we get:
- date / time of both departure and arrival
- location of both places
- distance of the movement
- time spent moving
- Cardinal direction

Most of those are relevant for the common movement studies, some metrics more than others. To do this, we will use dplyr.

We shall group by our desired data (tag_id, time, locations etc.), and use dplyr's lead / lag, to count between two subsequent rows in a tidy table, which finds the first detection, then connects it to the next. The rest, simple.

```{r}
dat1 <- dat %>% 
  transmute(tag_id = transmitter_id,
            datetime = detection_datetime, 
            station_name = receiver_name,
            latitude = receiver_deployment_latitude,
            longitude = receiver_deployment_longitude,
            sex = animal_sex) %>% 
  mutate(tag_id = as.character(tag_id)) %>% 
  filter(if_all(everything(), ~ !is.na(.))) # remove NAs and their entire row

anyNA(dat1) # needs to be false
str(dat1)
# format should be: 
# tag id = character
# datetime = POSIXct
# location = character
# station name = chr
# lat / lon = numeric
# sex = chr

# non-residency -----------------------------------------------------------

move_base <- dat1 %>%
  arrange(tag_id, datetime) %>% # arrange by ID and time
  group_by(tag_id) %>% # process data by tag ID, don't mix IDs
  mutate(next_time = lead(datetime), # lead time (lead / lag connects two rows (detections) that have already been arranged by time)
    next_station_name = lead(station_name), # lead location name, can be changed to receiver
    next_latitude = lead(latitude), # lat 
    next_longitude = lead(longitude)) %>% # lon
  # Define a movement as a change in location
  filter(!is.na(next_station_name), station_name != next_station_name) %>% # filter movements that go to the same location (my methods but you can remove)
  mutate(movement_id = row_number()) %>% # movement_id to connect an arrival / departure movement as one in future 
  ungroup() 

# lead created data for detecion 2, and replicated all variables
# we now use this dataframe to create arrival and departure data
# departures are detection 1, arrivals are detection 2 in a 'non-residency' or 'movement'
# we then use bind_rows to connect them 
# the output gives us one row per arrival / departure, with all data we need

# Departure rows creation
departures <- move_base %>%
  transmute(tag_id, # transmute brings over data, arranges in order, and keeps str as well as renames all in one
            datetime = datetime, # first datetime val in the dataframe
            sex = sex,
            station_name = station_name,
            latitude = latitude,
            longitude = longitude,
            movement = "departure",
            movement_id)

# Arrival rows
arrivals <- move_base %>%
  transmute(tag_id, # transmute brings over data, arranges in order, and keeps str as well as renames all in one
            datetime = next_time,
            sex = sex,
            station_name = next_station_name,
            latitude = next_latitude,
            longitude = next_longitude,
            movement = "arrival",
            movement_id)

# Combine
dat2 <- bind_rows(departures, arrivals) %>% # departure and arrival dataframes should be the same size in obs (they're halfs of the same df movebase)
  arrange(tag_id, movement_id, movement, datetime)

str(dat2)
table(dat2$station_name)
# dat2 should be double the obs of movebase
# since all we did was split rows of a movement into two: arrivals and depatures

datxy <- dat2 %>%
  group_by(station_name, tag_id, latitude, longitude, movement) %>%
  summarise(num_det = n(), .groups = 'drop')

IMOSxy_sf <- sf::st_as_sf(datxy, coords = c("longitude",
                        "latitude"), crs = 4326, agr = "constant")

mapview::mapview(IMOSxy_sf, cex = "num_det", zcol = "movement", fbg = FALSE)


print(dat2, n = 30)
```

Great! That worked, now let's do residency. Residency has a lot of different meanings (https://link.springer.com/content/pdf/10.1186/s40462-022-00364-z.pdf), so make sure you read up on what you want to ask, and how you want to calculate that. I wanted to look at long-term residency, so I set minimum detections to be two a day, and a minimum of two days to categorise a residency 'event'.

The way we do this, is very simple and again with dplyr. We will use the functions group_by and arrange to again format our data as we wish, then an if_else statement within mutate let's us see if minimum numbers of our thresholds have been met, if they have, we make new columns for our 'residency'! We also use functions like filter, summarise,  and select to clean up our outputs and present them nicely.

```{r}

min_detections_per_day <- 2   # minimum detections required on each day
min_res_days           <- 2   # minimum length of residency in days (calendar, inclusive)
max_gap_secs           <- 60*60*24  # 1 day in seconds

# residency ------------------------------------------------------------

residency <- dat1 %>%
  arrange(tag_id, station_name, datetime) %>% 
  group_by(tag_id, station_name) %>%
  mutate(time_gap = as.numeric(difftime(datetime, lag(datetime), units = "secs")),
    new_event = ifelse(is.na(time_gap) | time_gap > max_gap_secs, 1L, 0L),
    event_id = cumsum(replace_na(new_event, 1L))) %>%
  ungroup() %>%
  group_by(tag_id, station_name, event_id) %>%
  # compute daily stats within each event
  mutate(day = as_date(datetime)) %>%
  summarise(start_datetime = min(datetime),
    end_datetime = max(datetime),
    start_day = min(day),
    end_day = max(day),
    n_days_incl = as.integer(end_day - start_day) + 1L,   # inclusive day span
    n_detections = n(),
    # per-day detection counts
    days_meeting_threshold = sum( tapply(rep(1, n()), day, length) >= min_detections_per_day),
    sex = first(sex),
    .groups = "drop_last") %>%
  # keep events that last long enough AND meet the per-day threshold on every day
  filter(n_days_incl >= min_res_days,
    days_meeting_threshold == n_days_incl) %>%
  ungroup() %>%
  arrange(tag_id, start_datetime) %>% 
  select(-c(n_days_incl))

unique(residency$station_name)

residency
table(residency$station_name)
```

Fantastic, we found two residency events longer than our minimum thresholds. This were two days each, one had four detections and one had five! You can tweak your input minimums based on what kinda 'residency' you want to define, and detect.

That's all I have for you now, it's late, and I want to sleep.

Goodnight friends!
P
