---
title: "Functional programming · Intermediate (purrr & furrr)"
description: "Intermediate track — twenty quests on map, safely, walk, and parallel furrr."
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Intermediate (purrr & furrr)

Functional programming maps **functions over structures** instead of hand-writing loops. **purrr** makes that explicit: every map returns a predictable object type — essential when your data are split by site, tag, year, or burst.

Lists show up constantly in ecology — receivers nested in arrays, burst segments, model outputs per transect. Short quests with immediate output mirror how working memory stays manageable during learning ([Bransford et al., 2000](https://nap.nationalacademies.org/catalog/9853/how-people-learn-brain-mind-experience-and-school-expanded-edition)). You will learn when **`map2()`** fits and when **`pmap()`** is the right tool.

Parallel **`furrr`** is optional horsepower after correctness. Get the serial map right, then scale — the same habit as validating one tag before running the whole array.

On dense quests (**`map`**, **`map2`**, **`pmap`**, **`safely`**, **`map_dfr`**), you will practice the same function on **two or three ecological tables**.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(ggplot2)
})
```

## Q1 · Lists as ecology objects {#q01}

An R **list** holds heterogeneous objects — data frames, vectors, metadata — in one named container. Telemetry projects often store receiver info and detection tables as list elements before flattening.

Lists are why AI code suddenly returns `[[1]]` errors: you skipped a level. Building lists manually teaches inspection with `str()`.

Nest reef data frames by site as list elements.

```{r q01}
by_site <- split(reef, reef$site)                   # base split to named list
class(by_site)                                          # should be 'list'
names(by_site)                                          # site names as list names
lapply(by_site, nrow)                                     # rows per site (map() in Q2)
```

## Q2 · map() intuition {#q02}

**`map()`** applies a function to each list element and returns a list of results — always, unlike the `sapply()` simplification surprises in base R.

Think ‘for each element, do FUN’ — same loop pattern, fewer index typos when AI drafts batch jobs.

Extract unique species names per site using map.

```{r q02}
species_lists <- map(by_site, ~ unique(.x$species))  # .x is each site df
species_lists                                           # list of character vectors
lengths(species_lists)                                  # how many species per site
```

**Example 2 — fishery ports:** **`split`** landings by **`port`**, **`map`** **`nrow`**.

```{r q02-fishery}
by_port <- split(fishery, fishery$port)
map_int(by_port, nrow)
```

**Example 3 — acoustic zones:** **`map_chr`** the dominant **`receiver`** per zone (first row demo).

```{r q02-acoustic}
by_zone <- split(acoustic, acoustic$reef_zone)
map_chr(by_zone, ~ .x$receiver[1])
```

## Q3 · map_dbl(), map_chr() {#q03}

Typed maps **`map_dbl()`**, **`map_int()`**, **`map_chr()`** assert return types — the purrr analogue of `vapply()` from Level 1.

Use typed maps when building numeric summaries from lists before ggplot.

Mean fish count per site as a named numeric vector.

```{r q03}
mean_counts <- map_dbl(by_site, ~ mean(.x$count))   # numeric vector out
mean_counts                                             # named by site
sum(mean_counts)                                        # sanity check total
```

## Q4 · map2() paired columns {#q04}

**`map2()`** walks two parallel vectors element-wise — year paired with species, depth paired with pitch.

If lengths differ, purrr recycles with rules — verify `length(x) == length(y)` before trusting AI-generated map2.

Pair fishery year and species into composite labels.

```{r q04}
labels <- map2_chr(fishery$year, fishery$species, ~ paste(.x, .y, sep = "-"))
head(labels)                                            # first few composite ids
length(labels) == nrow(fishery)                         # lengths must match
```

**Example 2 — reef:** **`map2_dbl`** on **`count`** and **`depth_m`** for count-to-depth ratio.

```{r q04-reef}
map2_dbl(reef$count, reef$depth_m, function(c, d) if (d > 0) c / d else NA_real_)[1:5]
```

**Example 3 — biologging:** **`map2_dbl`** on consecutive depths within burst 1 (step changes).

```{r q04-biolog}
b1 <- biolog[biolog$burst_id == 1, ]
map2_dbl(b1$depth_m[-length(b1$depth_m)], b1$depth_m[-1], ~ .y - .x)[1:5]
```

## Q5 · pmap() row-wise {#q05}

**`pmap()`** maps over rows of a data frame — row-wise logic without an explicit index `i`.

Prefer vectorised `mutate()` when possible; pmap shines when calling external functions per row.

Build text summaries for the first five acoustic rows.

```{r q05}
ac5 <- acoustic |> dplyr::slice_head(n = 5)          # tiny slice
summaries <- pmap_chr(ac5, function(animal_id, receiver, detections, reef_zone) {
  paste0(animal_id, " @ ", receiver, ": ", detections, " (", reef_zone, ")")
})
summaries                                               # one string per row
```

**Example 2 — reef:** **`pmap_chr`** on first three transect rows ( **`transect_id`**, **`species`**, **`count`** ).

```{r q05-reef}
r3 <- reef |> slice_head(n = 3)
pmap_chr(r3, function(transect_id, site, species, count, depth_m) {
  paste0(transect_id, " ", species, ": ", count, " @ ", site)
})
```

**Example 3 — fishery:** **`pmap`** on **`year`**, **`species`**, **`tonnes`** for report lines.

```{r q05-fishery}
f3 <- fishery |> slice_head(n = 3)
pmap_chr(f3, function(year, species, tonnes, port) paste0(year, " ", species, " ", tonnes, "t @ ", port))
```

## Q6 · map_dfr() simulations {#q06}

**`map_dfr()`** row-binds data frame results — ideal for simulation replicates or per-group outputs.

Each iteration must return compatible columns; bind fails loudly otherwise — good QC for AI loops.

Simulate three Poisson draws per reef species name.

```{r q06}
set.seed(7)                                       # reproducible draws
sp <- unique(reef$species)                              # species labels
sim <- map_dfr(sp, function(s) {                        # one df per species
  tibble(species = s, draw = rpois(3, lambda = 5))      # three random counts
})
sim                                                     # stacked draws
```

**Example 2 — fishery:** three **`runif`** landings per **`port`**.

```{r q06-fishery}
ports <- unique(fishery$port)
map_dfr(ports, function(p) {
  tibble(port = p, tonnes = runif(3, 20, 90))
})
```

**Example 3 — acoustic:** simulate **`rpois`** detection counts per **`receiver`**.

```{r q06-acoustic}
rx <- unique(acoustic$receiver)[1:4]
map_dfr(rx, function(r) tibble(receiver = r, sim_det = rpois(3, lambda = 4)))
```

## Q7 · safely() errors as data {#q07}

**`safely()`** wraps a function to return `result` or `error` instead of stopping — essential on messy field spreadsheets.

Turning errors into data lets you filter failed rows and fix upstream instead of silent NA cascades.

Safely parse numeric strings including a bad value.

```{r q07}
raw <- c("12", "4.5", "NA", "oops")               # mixed strings
parsed <- map(raw, safely(as.numeric))                  # list of result/error
ok <- map_lgl(parsed, ~ is.null(.x$error))              # which succeeded
map_dbl(parsed[ok], "result")                           # extract numeric successes
```

**Example 2 — reef counts as character** (typo **`"12a"`** in a fake vector).

```{r q07-reef}
raw_counts <- c("8", "12", "12a", "3")
map(raw_counts, safely(as.numeric))
```

**Example 3 — fishery tonnes strings** with empty field.

```{r q07-fishery}
raw_t <- c("45.2", "", "88")
map(raw_t, safely(as.numeric))
```

## Q8 · walk() side effects {#q08}

**`walk()`** maps for side effects such as messages or file writes and discards return values.

Use walk to log QC per site without collecting useless NULLs in a list.

Print simple site headers for reef splits.

```{r q08}
walk(names(by_site), function(nm) {               # iterate site names
  message("Site: ", nm, " — rows: ", nrow(by_site[[nm]]))  # logging side effect
})
invisible(NULL)                                         # walk returns input invisibly
```

## Q9 · reduce() combine lists {#q09}

**`reduce()`** folds a list with a binary function — combine vectors or merge tables stepwise.

Light introduction: sum detection vectors conceptually before bind-heavy pipelines.

Sum all detection values via reduce on split vectors.

```{r q09}
by_an <- split(acoustic$detections, acoustic$animal_id)  # list of nums
per_an <- map_dbl(by_an, sum)                           # one total per animal tag
total_det <- reduce(per_an, `+`)                        # fold scalar sums together
total_det                                               # grand total detections
```

## Q10 · list-columns intro {#q10}

Tibbles can store **list-columns** — each cell holds a vector or mini data frame before `unnest()`.

AI-generated nested JSON often becomes list-columns; recognise them in `glimpse()` output.

Nest biolog seconds and depths per burst_id.

```{r q10}
nested_bio <- biolog |>                           # biologging table
  group_by(burst_id) |>                                 # one row per burst soon
  summarise(times = list(seconds), depths = list(depth_m), .groups = "drop")
nested_bio                                              # list-columns visible
nested_bio$depths[[1]]                                  # first burst depth vector
```

## Q11 · map on telemetry splits {#q11}

Split acoustic data by **`animal_id`**, map summaries — the animal-centric pattern in movement ecology.

Check empty animals: map should still return sensible values or explicit NA.

Maximum detections per animal as a named vector.

```{r q11}
by_animal <- group_split(acoustic, animal_id)   # list of tibbles
max_det <- map_dbl(by_animal, ~ max(.x$detections))    # peak ping count
names(max_det) <- map_chr(by_animal, ~ unique(.x$animal_id))  # name elements
max_det                                                 # per-tag max
```

## Q12 · map on fishery years {#q12}

Split **`fishery`** by year and map total tonnes — functional annual report tables.

Named results from map reduce join friction when plotting.

Total tonnes landed each calendar year.

```{r q12}
by_year <- group_split(fishery, year)            # one df per year
year_tot <- map_dbl(by_year, ~ sum(.x$tonnes))           # annual totals
setNames(year_tot, map_int(by_year, ~ unique(.x$year)))  # names = years
year_tot                                                # numeric summary
```

## Q13 · map2 weights & counts {#q13}

**`map2()`** on paired vectors supports weighted calculations — tonnes times price, detections times effort.

Always verify equal lengths on real tagging projects before map2.

Element-wise product of depth and unit weights (demo).

```{r q13}
x <- biolog$depth_m[1:10]                         # depths subsample
w <- rep(1, 10)                                         # equal weights demo
weighted <- map2_dbl(x, w, ~ .x * .y)                    # element products
sum(weighted) / sum(w)                                  # weighted mean depth
```

## Q14 · possibly() softer failures {#q14}

**`possibly()`** replaces errors with a default value — softer than `safely()` when you want NA back.

Do not hide real bugs in production; use for known messy sensor drops only.

Possibly divide detections by multipliers with a zero divisor.

```{r q14}
mult <- c(1, 2, 0, 4)                              # includes zero divisor
safe_div <- possibly(function(a, b) a / b, otherwise = NA_real_)
map2_dbl(acoustic$detections[1:4], mult, safe_div)    # NA where divide fails
```

## Q15 · quietly() hide messages {#q15}

**`quietly()`** captures messages while returning results — useful when mapped functions are chatty.

Keeps R Markdown clean while preserving access to logs via the result object.

Quietly compute means on site count vectors.

```{r q15}
chatty_mean <- function(x) {                      # function with message
  message("computing mean on length ", length(x))
  mean(x)
}
q <- quietly(chatty_mean)                               # wrap to capture noise
res <- map(by_site, ~ q(.x$count))                      # list of quiet results
map_dbl(res, "result")                                  # extract means only
```

## Q16 · furrr parallel map {#q16}

**furrr** **`future_map()`** mirrors **`map()`** on parallel workers — worth it on heavy simulations, not tiny reef tables.

Install furrr/future if missing; serial map remains the correctness baseline.

Optional parallel mean counts per site when furrr is installed.

```{r q16}
has_furrr <- requireNamespace("furrr", quietly = TRUE)  # check package
if (has_furrr) {
  suppressPackageStartupMessages({ library(furrr); library(future) })
  plan(multisession, workers = 2)                       # local worker pool
  future_map_dbl(by_site, ~ mean(.x$count))            # parallel map
} else {
  message("Install furrr for parallel demo; using map_dbl")
  map_dbl(by_site, ~ mean(.x$count))                   # serial fallback
}
```

## Q17 · future plans & workers {#q17}

The **future** **`plan()`** selects sequential vs multisession execution — set once before furrr maps.

More workers add overhead; small tables rarely benefit — reason about cost like any field campaign.

Time serial map over site list as baseline.

```{r q17}
time_serial <- system.time({                       # time serial map
  map_dbl(by_site, ~ mean(.x$count))
})
time_serial                                             # print timing object
```

## Q18 · map deep on biologging {#q18}

Map over burst splits to build hierarchical summaries — ‘deep’ mapping across nested structure.

Sketch the list tree before accepting AI nested map code.

Per-burst mean depth and record count bound into one table.

```{r q18}
burst_stats <- biolog |>                         # biologging data
  group_split(burst_id) |>                              # list of burst tibbles
  map(~ summarise(.x, mean_depth = mean(depth_m), n_sec = n()))
burst_tbl <- map_dfr(burst_stats, identity)             # row-bind summaries
head(burst_tbl)                                         # one row per burst
```

## Q19 · compose functions {#q19}

**`compose()`** builds function pipelines — conceptual kin to `|>` but for function values passed to map.

Useful for reusable QC plus transform blocks in functional workflows.

Compose rounding after min-max scaling of fishery tonnes.

```{r q19}
scale01 <- function(x) (x - min(x)) / (max(x) - min(x))  # min-max scale
round2 <- function(x) round(x, 2)                       # two decimals
fmt <- compose(round2, scale01)                         # combined function
fmt(fishery$tonnes)                                     # formatted vector
```

## Q20 · purrr capstone {#q20}

Integrate split, map, and `map_dfr()` on acoustic data — the workflow to verify before batch-processing receiver exports with AI.

Inspect list depth at each step; capstones fail when an extra `[[1]]` sneaks in.

Zone summaries bound and plotted.

```{r q20}
zone_list <- group_split(acoustic, reef_zone)    # list by habitat zone
zone_summary <- map_dfr(zone_list, function(df) {
  tibble(
    reef_zone = unique(df$reef_zone),
    total_det = sum(df$detections),
    n_rows = nrow(df)
  )
})
ggplot(zone_summary, aes(x = reef_zone, y = total_det, fill = reef_zone)) +
  geom_col(show.legend = FALSE) +                       # bar chart by zone
  labs(title = "Detections by reef zone", y = "Sum of detections")
```

