---
title: "Functional programming · Intermediate (purrr & furrr)"
description: "Intermediate · purrr"
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Intermediate (purrr & furrr)
Lists and `map` for tags, bursts, years. Do Base R + dplyr first.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(ggplot2)
})
```

## Q1 · Lists as ecology objects {#q01}
```{r q01}
by_site <- split(reef, reef$site) # base split to named list
class(by_site) # should be 'list'
names(by_site) # site names as list names
lapply(by_site, nrow) # rows per site
```

## Q2 · map() intuition {#q02}
```{r q02}
species_lists <- map(by_site, ~ unique(.x$species)) # .x is each site df
species_lists # list of character vectors
lengths(species_lists) # how many species per site
```

**Ex 2.**

```{r q02-fishery}
by_port <- split(fishery, fishery$port)
map_int(by_port, nrow)
```

**Ex 2.**

```{r q02-acoustic}
by_zone <- split(acoustic, acoustic$reef_zone)
map_chr(by_zone, ~ .x$receiver[1])
```

## Q3 · map_dbl(), map_chr() {#q03}
```{r q03}
mean_counts <- map_dbl(by_site, ~ mean(.x$count)) # numeric vector out
mean_counts # named by site
sum(mean_counts) # sanity check total
```

## Q4 · map2() paired columns {#q04}
```{r q04}
labels <- map2_chr(fishery$year, fishery$species, ~ paste(.x, .y, sep = "-"))
head(labels)
length(labels) == nrow(fishery) # lengths must match
```

**Ex 2.**

```{r q04-reef}
map2_dbl(reef$count, reef$depth_m, function(c, d) if (d > 0) c / d else NA_real_)[1:5]
```

**Ex 2.**

```{r q04-biolog}
b1 <- biolog[biolog$burst_id == 1, ]
map2_dbl(b1$depth_m[-length(b1$depth_m)], b1$depth_m[-1], ~ .y - .x)[1:5]
```

## Q5 · pmap() row-wise {#q05}
```{r q05}
ac5 <- acoustic |> dplyr::slice_head(n = 5) # tiny slice
summaries <- pmap_chr(ac5, function(animal_id, receiver, detections, reef_zone) {
  paste0(animal_id, " @ ", receiver, ": ", detections, " (", reef_zone, ")")
})
summaries # one string per row
```

**Ex 2.**

```{r q05-reef}
r3 <- reef |> slice_head(n = 3)
pmap_chr(r3, function(transect_id, site, species, count, depth_m) {
  paste0(transect_id, " ", species, ": ", count, " @ ", site)
})
```

**Ex 2.**

```{r q05-fishery}
f3 <- fishery |> slice_head(n = 3)
pmap_chr(f3, function(year, species, tonnes, port) paste0(year, " ", species, " ", tonnes, "t @ ", port))
```

## Q6 · map_dfr() simulations {#q06}
```{r q06}
set.seed(7) # reproducible draws
sp <- unique(reef$species) # species labels
sim <- map_dfr(sp, function(s) {
  tibble(species = s, draw = rpois(3, lambda = 5)) # three random counts
})
sim # stacked draws
```

**Ex 2.**

```{r q06-fishery}
ports <- unique(fishery$port)
map_dfr(ports, function(p) {
  tibble(port = p, tonnes = runif(3, 20, 90))
})
```

**Ex 2.**

```{r q06-acoustic}
rx <- unique(acoustic$receiver)[1:4]
map_dfr(rx, function(r) tibble(receiver = r, sim_det = rpois(3, lambda = 4)))
```

## Q7 · safely() errors as data {#q07}
```{r q07}
raw <- c("12", "4.5", "NA", "oops") # mixed strings
parsed <- map(raw, safely(as.numeric)) # list of result/error
ok <- map_lgl(parsed, ~ is.null(.x$error)) # which succeeded
map_dbl(parsed[ok], "result") # extract numeric successes
```

**Ex 2.**

```{r q07-reef}
raw_counts <- c("8", "12", "12a", "3")
map(raw_counts, safely(as.numeric))
```

**Ex 2.**

```{r q07-fishery}
raw_t <- c("45.2", "", "88")
map(raw_t, safely(as.numeric))
```

## Q8 · walk() side effects {#q08}
```{r q08}
walk(names(by_site), function(nm) { # iterate site names
  message("Site: ", nm, " — rows: ", nrow(by_site[[nm]]))
})
invisible(NULL) # walk returns input invisibly
```

## Q9 · reduce() combine lists {#q09}
```{r q09}
by_an <- split(acoustic$detections, acoustic$animal_id) # list of nums
per_an <- map_dbl(by_an, sum) # one total per animal tag
total_det <- reduce(per_an, `+`) # fold scalar sums together
total_det # grand total detections
```

## Q10 · list-columns intro {#q10}
```{r q10}
nested_bio <- biolog |> # biologging table
  group_by(burst_id) |>
  summarise(times = list(seconds), depths = list(depth_m), .groups = "drop")
nested_bio # list-columns visible
nested_bio$depths[[1]]
```

## Q11 · map on telemetry splits {#q11}
```{r q11}
by_animal <- group_split(acoustic, animal_id) # list of tibbles
max_det <- map_dbl(by_animal, ~ max(.x$detections)) # peak ping count
names(max_det) <- map_chr(by_animal, ~ unique(.x$animal_id)) # name elements
max_det # per-tag max
```

## Q12 · map on fishery years {#q12}
```{r q12}
by_year <- group_split(fishery, year)
year_tot <- map_dbl(by_year, ~ sum(.x$tonnes)) # annual totals
setNames(year_tot, map_int(by_year, ~ unique(.x$year))) # names = years
year_tot # numeric summary
```

## Q13 · map2 weights & counts {#q13}
```{r q13}
x <- biolog$depth_m[1:10] # depths subsample
w <- rep(1, 10) # equal weights demo
weighted <- map2_dbl(x, w, ~ .x * .y) # element products
sum(weighted) / sum(w) # weighted mean depth
```

## Q14 · possibly() softer failures {#q14}
```{r q14}
mult <- c(1, 2, 0, 4) # includes zero divisor
safe_div <- possibly(function(a, b) a / b, otherwise = NA_real_)
map2_dbl(acoustic$detections[1:4], mult, safe_div) # NA where divide fails
```

## Q15 · quietly() hide messages {#q15}
```{r q15}
chatty_mean <- function(x) { # function with message
  message("computing mean on length ", length(x))
  mean(x)
}
q <- quietly(chatty_mean) # wrap to capture noise
res <- map(by_site, ~ q(.x$count)) # list of quiet results
map_dbl(res, "result")
```

## Q16 · furrr parallel map {#q16}
```{r q16}
has_furrr <- requireNamespace("furrr", quietly = TRUE) # check package
if (has_furrr) {
  suppressPackageStartupMessages({ library(furrr); library(future) })
  plan(multisession, workers = 2) # local worker pool
  future_map_dbl(by_site, ~ mean(.x$count)) # parallel map
} else {
  message("Install furrr for parallel demo; using map_dbl")
  map_dbl(by_site, ~ mean(.x$count)) # serial fallback
}
```

## Q17 · future plans & workers {#q17}
```{r q17}
time_serial <- system.time({ # time serial map
  map_dbl(by_site, ~ mean(.x$count))
})
time_serial # print timing object
```

## Q18 · map deep on biologging {#q18}
```{r q18}
burst_stats <- biolog |>
  group_split(burst_id) |> # list of burst tibbles
  map(~ summarise(.x, mean_depth = mean(depth_m), n_sec = n()))
burst_tbl <- map_dfr(burst_stats, identity) # row-bind summaries
head(burst_tbl)
```

## Q19 · compose functions {#q19}
```{r q19}
scale01 <- function(x) (x - min(x)) / (max(x) - min(x)) # min-max scale
round2 <- function(x) round(x, 2) # two decimals
fmt <- compose(round2, scale01)
fmt(fishery$tonnes)
```

## Q20 · purrr capstone {#q20}
```{r q20}
zone_list <- group_split(acoustic, reef_zone) # list by habitat zone
zone_summary <- map_dfr(zone_list, function(df) {
  tibble(
    reef_zone = unique(df$reef_zone),
    total_det = sum(df$detections),
    n_rows = nrow(df)
  )
})
ggplot(zone_summary, aes(x = reef_zone, y = total_det, fill = reef_zone)) +
  geom_col(show.legend = FALSE) + # bar chart by zone
  labs(title = "Detections by reef zone", y = "Sum of detections")
```

