---
title: "Functional programming · Intermediate (dplyr)"
description: "Intermediate track — twenty quests on mutate, group_by, lead/lag, joins, and pivots."
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Intermediate (dplyr)

**dplyr** verbs read like sentences: select columns, filter rows, mutate new variables, summarise within groups. This level builds verb-by-verb mental models so you can read, write, and fix pipelines on reef, fishery, acoustic, and biologging data.

Learning progresses when you connect new syntax to problems you already solved in base R ([Bransford et al., 2000](https://nap.nationalacademies.org/catalog/9853/how-people-learn-brain-mind-experience-and-school-expanded-edition)). Each quest (~10 minutes) ends with a prediction you can check: row counts, column names, or a plot you recognise from the field.

Treat **dplyr** as a contract with your data: every verb documents an intent. Choosing which steps belong in a pipeline is the same scientific reasoning habit Klahr and Dunbar describe in experiment design — applied to code.

Harder verbs (**`group_by`**, **`lead`/`lag`**, **`case_when`**, **`across`**, joins, pivots) include **multiple worked examples** on different ecological tables.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
})
```

## Q1 · select() & rename() {#q01}

The **`select()`** verb keeps or drops columns by name, type, or tidy helpers like `starts_with()`. **`rename()`** changes names without touching other columns — cleaner than reassigning `names(df)` when AI adds a typo.

Before typing `select()`, list which columns your analysis actually needs. Extra columns increase cognitive load when reading joins downstream.

We trim the reef table to identifiers and abundance, renaming `count` to `n_fish` for publication-ready headers.

```{r q01}
reef_sl <- reef |>                                 # start with reef transect data
  select(transect_id, site, species, count, depth_m) |>  # keep core columns only
  rename(n_fish = count)                               # clearer name for counts
glimpse(reef_sl)                                       # confirm column names
```

## Q2 · filter() rows {#q02}

**`filter()`** keeps rows where a logical condition is TRUE. Unlike base subsetting, dplyr verbs error helpfully when you use `=` instead of `==` in many contexts — but AI still generates wrong variable names.

Always check `nrow()` before and after a filter; it is cheap feedback on whether your logic matches the field protocol.

Keep East site transects with depth at least five metres.

```{r q02}
reef_east_deep <- reef |>                        # pipe data into filter
  filter(site == "East", depth_m >= 5)                  # both conditions must hold
nrow(reef_east_deep)                                    # rows retained
distinct(reef_east_deep, transect_id)                   # which transects survived
```

## Q3 · mutate() new columns {#q03}

**`mutate()`** adds or replaces columns while keeping all rows — the workhorse for unit conversions, binary flags, and intermediate biology.

Mutate is vectorised: each expression sees whole columns. That matches how R thinks and differs from row-wise loops unless you call `rowwise()`.

Add kg from tonnes on fishery landings and a heavy-landing flag.

```{r q03}
fishery_mut <- fishery |>                        # fisheries landing table
  mutate(
    kg = tonnes * 1000,                                 # metric conversion
    heavy = tonnes > 80                                 # logical flag for large trips
  )
summary(fishery_mut$kg)                                 # QC numeric range
```

## Q4 · transmute() keep only new {#q04}

**`transmute()`** is mutate that drops all previous columns — useful when you only want derived variables for modelling or export.

AI sometimes chains `select()` after `mutate()` when `transmute()` would be clearer; recognising the verb saves clutter.

Extract animal id and scaled detections from acoustic data.

```{r q04}
acoustic_derived <- acoustic |>                  # acoustic telemetry rows
  transmute(
    animal_id,                                          # keep tag id
    detections_scaled = detections / max(detections)  # scale to 0-1 for plotting
  )
head(acoustic_derived)                                  # only new columns remain
```

## Q5 · arrange() sort rows {#q05}

**`arrange()`** sorts rows using `order()` logic under the hood — including `desc()` for decreasing sort.

Sorting is not cosmetic: time-series plots and `lead()`/`lag()` in later quests assume intentional row order within groups.

Order biologging seconds within each burst.

```{r q05}
biolog_ord <- biolog |>                          # biologging burst series
  arrange(burst_id, seconds)                            # time order inside bursts
head(biolog_ord, 6)                                     # verify first burst sequence
```

## Q6 · distinct() unique rows {#q06}

**`distinct()`** returns unique rows, optionally by a subset of columns — dplyr’s answer to `unique()` on data frames.

Use it after joins to confirm you did not duplicate keys — a common AI join mistake.

List unique site–depth combinations on reef surveys.

```{r q06}
sites_depths <- reef |>                          # reef data
  distinct(site, depth_m) |>                            # one row per combo
  arrange(site, depth_m)                                # readable table
sites_depths                                            # print result
```

## Q7 · slice() row index {#q07}

**`slice()`** picks rows by position; helpers like **`slice_max()`** pick top values per group when combined with `group_by()`.

Index slicing is fragile if row order changes — document why order is stable before slicing.

Take the deepest reading per burst from biologging.

```{r q07}
deepest <- biolog |>                             # all burst seconds
  group_by(burst_id) |>                                 # partition by burst
  slice_max(depth_m, n = 1, with_ties = FALSE)          # deepest single row each
deepest                                                 # one row per burst_id
```

## Q8 · group_by() + summarise() {#q08}

**`group_by()`** marks grouping columns; **`summarise()`** collapses each group to one row. Forgetting to ungroup is a classic AI bug that poisons later mutates.

After summarise, ask: ‘How many rows should I have?’ — compare to `n_distinct(group_cols)`.

Mean fish count per site on reef data.

```{r q08}
reef_site_mean <- reef |>                        # transect counts
  group_by(site) |>                                     # groups: North, South, East
  summarise(
    mean_count = mean(count),                           # average abundance
    n_rows = n(),                                       # rows contributing
    .groups = "drop"                                    # avoid sticky grouping
  )
reef_site_mean                                          # three rows expected
```

**Example 2 — fishery:** mean **`tonnes`** landed per **`species`**.

```{r q08-fishery}
fishery |>
  group_by(species) |>
  summarise(mean_t = mean(tonnes), n = n(), .groups = "drop")
```

**Example 3 — acoustic:** total **`detections`** per **`animal_id`** (sum, not mean — common telemetry summary).

```{r q08-acoustic}
acoustic |>
  group_by(animal_id) |>
  summarise(total_det = sum(detections), n_receivers = n_distinct(receiver), .groups = "drop")
```

## Q9 · count() quick tables {#q09}

**`count()`** is shorthand for `group_by()` + `summarise(n = n())`, ideal for frequency tables before modelling.

Run counts before chi-square tests or occupancy models to spot empty cells early.

Count acoustic detections rows by reef zone and receiver.

```{r q09}
det_counts <- acoustic |>                        # telemetry table
  count(reef_zone, receiver, name = "rows", sort = TRUE)  # sorted frequencies
head(det_counts, 10)                                    # busiest combinations
```

## Q10 · lead() & lag() time series {#q10}

**`lag()`** shifts values backward in row order; **`lead()`** shifts forward — essential for step changes in depth or pitch along `seconds`.

They require sorted data within groups — you arranged biologging in Q5 for this reason.

Compute second-to-second depth change within bursts.

```{r q10}
depth_steps <- biolog_ord |>                     # sorted biolog data
  group_by(burst_id) |>                                 # independent time within burst
  mutate(depth_change = depth_m - lag(depth_m)) |>    # current minus previous
  filter(!is.na(depth_change))                          # drop first second each burst
head(depth_steps, 8)                                    # inspect changes
```

**Example 2 — biologging pitch:** same pattern on **`pitch_deg`** within bursts.

```{r q10-pitch}
biolog_ord |>
  group_by(burst_id) |>
  mutate(pitch_step = pitch_deg - lag(pitch_deg)) |>
  filter(!is.na(pitch_step)) |>
  slice_head(n = 4)
```

**Example 3 — fishery (file order):** **`lag(tonnes)`** within **`species`** to compare consecutive landing records in the CSV order (not true calendar time — always check sort first).

```{r q10-fishery}
fishery |>
  arrange(species, year) |>
  group_by(species) |>
  mutate(tonnes_prev = lag(tonnes), delta_t = tonnes - lag(tonnes)) |>
  filter(!is.na(delta_t)) |>
  head(6)
```

## Q11 · if_else() & case_when() {#q11}

**`if_else()`** is strict typed ifelse for two branches; **`case_when()`** handles multiple rules readable top-to-bottom — like nested QC without spaghetti.

Order matters in `case_when()`: first match wins, mirroring priority QC protocols.

Classify fishery landings into size bins for reporting.

```{r q11}
fishery_bins <- fishery |>                       # landing records
  mutate(
    size_bin = case_when(
      tonnes < 30 ~ "small",                            # first matching rule
      tonnes < 70 ~ "medium",
      TRUE ~ "large"                                    # catch-all required
    )
  )
count(fishery_bins, size_bin)                           # bin frequencies
```

**Example 2 — reef depth zones** with **`case_when()`** (MPA planning bins).

```{r q11-reef}
reef |>
  mutate(depth_zone = case_when(
    depth_m < 5 ~ "upper",
    depth_m < 10 ~ "mid",
    TRUE ~ "deep"
  )) |>
  count(site, depth_zone)
```

**Example 3 — acoustic activity** with strict **`if_else()`** (two outcomes only).

```{r q11-acoustic}
acoustic |>
  mutate(active = if_else(detections >= 4, "high", "low")) |>
  count(reef_zone, active)
```

## Q12 · across() many columns {#q12}

**`across()`** applies a function to multiple columns selected by name or helper — the tidy way to scale numeric columns without copy-paste.

Watch types: `across(where(is.numeric))` avoids accidentally rounding character columns AI mislabeled.

Z-score numeric columns in biologging (within full table demo).

```{r q12}
biolog_scaled <- biolog |>                       # burst time series
  mutate(across(c(depth_m, pitch_deg), ~ as.numeric(scale(.))))  # z-scores
summary(biolog_scaled$depth_m)                          # mean ~0, sd ~1
```

**Example 2 — reef:** scale **`count`** and **`depth_m`** after filtering **`site == "North"`**.

```{r q12-reef}
reef |>
  filter(site == "North") |>
  mutate(across(c(count, depth_m), ~ as.numeric(scale(.)))) |>
  summarise(across(c(count, depth_m), mean))
```

**Example 3 — fishery:** **`across`** with **`mean`** and **`sd`** on **`tonnes`** by **`port`**.

```{r q12-fishery}
fishery |>
  group_by(port) |>
  summarise(across(tonnes, list(mean = mean, sd = sd), .names = "{.fn}_{.col}"))
```

## Q13 · group_modify() custom tables {#q13}

**`group_modify()`** passes each group as a tibble to a function that returns a tibble — when `summarise()` is too limited for custom row-wise tables.

Use sparingly; it is slower than vectorised summaries but mirrors base `lapply` on splits explicitly.

Per species on reef: min, max, and range of counts.

```{r q13}
sp_range <- reef |>                             # species counts
  group_by(species) |>                                  # one group per species
  group_modify(~ {                                      # .x is species subset
    tibble(                                             # return tibble shape
      min_n = min(.x$count),
      max_n = max(.x$count),
      range_n = max(.x$count) - min(.x$count)
    )
  })
sp_range                                                # custom table per species
```

**Example 2 — fishery:** range of **`tonnes`** per **`port`**.

```{r q13-fishery}
fishery |>
  group_by(port) |>
  group_modify(~ tibble(min_t = min(.x$tonnes), max_t = max(.x$tonnes)))
```

**Example 3 — acoustic:** top **`receiver`** by summed **`detections`** per **`animal_id`**.

```{r q13-acoustic}
acoustic |>
  group_by(animal_id) |>
  group_modify(~ {
    .x |>
      group_by(receiver) |>
      summarise(d = sum(detections), .groups = "drop") |>
      slice_max(d, n = 1)
  })
```

## Q14 · mutate(.by = ) {#q14}

Dplyr 1.1+ **`mutate(.by = )`** groups temporarily for one mutate without sticky `group_by()` — great for within-animal or within-burst calculations.

It reduces AI-generated `group_by()` + `ungroup()` boilerplate when only one verb needs groups.

Centre detections within each animal id.

```{r q14}
acoustic_cent <- acoustic |>                     # detection rows
  mutate(
    det_cent = detections - mean(detections),           # overall mean if no .by
    det_by_an = detections - mean(detections), .by = animal_id  # within-tag mean
  )
head(acoustic_cent)                                     # compare columns
```

## Q15 · left_join() metadata {#q15}

**`left_join()`** keeps all rows from the left table and adds matching columns from the right — the join ecologists use for tagging metadata.

Always join on explicit keys; verify row count unchanged unless you expect duplication.

Attach a synthetic port region lookup to fishery landings.

```{r q15}
ports <- tibble(                                 # small lookup table
  port = c("Cairns", "Broome", "Mooloolaba"),
  region = c("GBR", "NW", "SEQ")
)
fishery_geo <- fishery |>                               # landings left side
  left_join(ports, by = "port")                         # add region column
count(fishery_geo, region)                              # rows per region
```

**Example 2 — acoustic:** join tag **`metadata`** (hypothetical deployment reef) onto detections.

```{r q15-acoustic}
tag_meta <- tibble(
  animal_id = c("A101", "A102", "A103"),
  tag_type = c("V16", "V16", "V13")
)
acoustic |>
  left_join(tag_meta, by = "animal_id") |>
  count(tag_type, reef_zone)
```

**Example 3 — reef:** join **`site`** management zone onto transects.

```{r q15-reef}
site_meta <- tibble(site = c("North", "South", "East"), zone = c("MPA", "Open", "MPA"))
reef |>
  left_join(site_meta, by = "site") |>
  distinct(site, zone)
```

## Q16 · bind_rows() stack surveys {#q16}

**`bind_rows()`** stacks compatible data frames — binding years or survey legs after harmonising column names.

Unlike `rbind()`, it handles missing columns by filling NA — know which behaviour you want before AI stacks mismatched CSVs.

Stack reef subset with itself renamed as pseudo replicate survey.

```{r q16}
reef_a <- reef |> filter(site == "North")       # first leg
reef_b <- reef |> filter(site == "South") |>             # second leg
  mutate(transect_id = paste0(transect_id, "_b"))       # avoid id collision
stacked <- bind_rows(
  mutate(reef_a, leg = "north"),
  mutate(reef_b, leg = "south")
)                                                     # tagged stack
nrow(stacked)                                           # combined rows
```

## Q17 · pivot_longer() {#q17}

**`pivot_longer()`** melts wide columns into key-value rows — standard before ggplot when measurements live in many columns.

Name the `names_to` and `values_to` columns deliberately so AI-generated plots have sensible axis labels.

Wide fishery: columns per species for one year slice (constructed) then pivot.

```{r q17}
wide_demo <- fishery |>                          # long landings
  filter(year == 2020) |>                               # single year
  select(species, tonnes) |>                              # minimal columns
  pivot_wider(names_from = species, values_from = tonnes)  # wide for exercise
long_demo <- wide_demo |>                               # now melt back
  pivot_longer(everything(), names_to = "species", values_to = "tonnes", values_drop_na = TRUE)
long_demo                                               # long form restored
```

**Example 2 — reef:** one transect wide by species, then long again for plotting.

```{r q17-reef}
reef |>
  filter(transect_id == "T01") |>
  select(species, count) |>
  pivot_wider(names_from = species, values_from = count) |>
  pivot_longer(everything(), names_to = "species", values_to = "count")
```

**Example 3 — biologging:** **`pivot_longer()`** on **`depth_m`** and **`pitch_deg`** for one burst (sensor comparison).

```{r q17-biolog}
biolog |>
  filter(burst_id == 1) |>
  select(seconds, depth_m, pitch_deg) |>
  pivot_longer(c(depth_m, pitch_deg), names_to = "sensor", values_to = "value") |>
  head(6)
```

## Q18 · pivot_wider() {#q18}

**`pivot_wider()`** spreads key-value pairs into columns — useful for occupancy matrices or wide export to Excel.

Watch for duplicate keys; pivot will error or aggregate depending on settings — check `values_fn`.

Acoustic: detections per animal-receiver pair as wide matrix.

```{r q18}
ac_wide <- acoustic |>                          # long detections
  group_by(animal_id, receiver) |>                      # unique pairs
  summarise(dets = sum(detections), .groups = "drop") |> # aggregate duplicates
  pivot_wider(names_from = receiver, values_from = dets, values_fill = 0)
ac_wide                                                 # wide receiver columns
```

## Q19 · replace_na() & coalesce() {#q19}

**`replace_na()`** swaps NA for a value; **`coalesce()`** picks the first non-NA from several columns — handy when merging sensor streams.

Document why NA existed: missing vs not applicable changes whether replace is scientific.

Replace NA in optional column after join simulation on reef depths.

```{r q19}
reef_na <- reef |>                              # original reef
  mutate(depth_m = if_else(species == "Grouper", NA_real_, depth_m))  # inject NA demo
reef_fix <- reef_na |>                                  # repair for plotting
  mutate(depth_m = replace_na(depth_m, median(reef$depth_m, na.rm = TRUE)))
sum(is.na(reef_fix$depth_m))                            # should be zero
```

## Q20 · dplyr capstone pipeline {#q20}

Capstone: one readable pipeline from raw landings to a figure — the shape of a methods supplement AI should not invent without your sign-off.

Read bottom-up: what object enters `ggplot()`, what one row represents, which filters happened.

Fishery totals by species and port with plot.

```{r q20}
cap <- fishery |>                               # start landings
  filter(year >= 2019) |>                               # recent years
  group_by(species, port) |>                            # reporting cells
  summarise(total_t = sum(tonnes), .groups = "drop") |> # summed tonnes
  arrange(desc(total_t))                                # largest first
ggplot(cap, aes(x = reorder(species, total_t), y = total_t, fill = port)) +
  geom_col(position = "dodge") +                        # compare ports
  coord_flip() +                                        # readable labels
  labs(title = "Landings by species and port (2019+)", x = NULL, y = "Tonnes")
```

