---
title: "Functional programming · Intermediate (dplyr)"
description: "Intermediate · dplyr"
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Intermediate (dplyr)
dplyr verbs on tidy ecology tables. One idea per quest.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
})
```

## Q1 · select() & rename() {#q01}
```{r q01}
reef_sl <- reef |>
  select(transect_id, site, species, count, depth_m) |>
  rename(n_fish = count) # clearer name for counts
glimpse(reef_sl) # confirm column names
```

## Q2 · filter() rows {#q02}
```{r q02}
reef_east_deep <- reef |> # pipe data into filter
  filter(site == "East", depth_m >= 5) # both conditions must hold
nrow(reef_east_deep) # rows retained
distinct(reef_east_deep, transect_id) # which transects survived
```

## Q3 · mutate() new columns {#q03}
```{r q03}
fishery_mut <- fishery |> # fisheries landing table
  mutate(
    kg = tonnes * 1000,
    heavy = tonnes > 80 # logical flag for large trips
  )
summary(fishery_mut$kg) # QC numeric range
```

## Q4 · transmute() keep only new {#q04}
```{r q04}
acoustic_derived <- acoustic |>
  transmute(
    animal_id, # keep tag id
    detections_scaled = detections / max(detections) # scale to 0-1 for plotting
  )
head(acoustic_derived)
```

## Q5 · arrange() sort rows {#q05}
```{r q05}
biolog_ord <- biolog |> # biologging burst series
  arrange(burst_id, seconds) # time order inside bursts
head(biolog_ord, 6) # verify first burst sequence
```

## Q6 · distinct() unique rows {#q06}
```{r q06}
sites_depths <- reef |>
  distinct(site, depth_m) |> # one row per combo
  arrange(site, depth_m) # readable table
sites_depths # print result
```

## Q7 · slice() row index {#q07}
```{r q07}
deepest <- biolog |> # all burst seconds
  group_by(burst_id) |>
  slice_max(depth_m, n = 1, with_ties = FALSE)
deepest # one row per burst_id
```

## Q8 · group_by() + summarise() {#q08}
```{r q08}
reef_site_mean <- reef |> # transect counts
  group_by(site) |> # groups: North, South, East
  summarise(
    mean_count = mean(count), # average abundance
    n_rows = n(), # rows contributing
    .groups = "drop" # avoid sticky grouping
  )
reef_site_mean # three rows expected
```

**Ex 2.**

```{r q08-fishery}
fishery |>
  group_by(species) |>
  summarise(mean_t = mean(tonnes), n = n(), .groups = "drop")
```

**Ex 2.**

```{r q08-acoustic}
acoustic |>
  group_by(animal_id) |>
  summarise(total_det = sum(detections), n_receivers = n_distinct(receiver), .groups = "drop")
```

## Q9 · count() quick tables {#q09}
```{r q09}
det_counts <- acoustic |> # telemetry table
  count(reef_zone, receiver, name = "rows", sort = TRUE) # sorted frequencies
head(det_counts, 10) # busiest combinations
```

## Q10 · lead() & lag() time series {#q10}
```{r q10}
depth_steps <- biolog_ord |>
  group_by(burst_id) |>
  mutate(depth_change = depth_m - lag(depth_m)) |> # current minus previous
  filter(!is.na(depth_change))
head(depth_steps, 8) # inspect changes
```

**Ex 2.**

```{r q10-pitch}
biolog_ord |>
  group_by(burst_id) |>
  mutate(pitch_step = pitch_deg - lag(pitch_deg)) |>
  filter(!is.na(pitch_step)) |>
  slice_head(n = 4)
```

**Ex 2.**

```{r q10-fishery}
fishery |>
  arrange(species, year) |>
  group_by(species) |>
  mutate(tonnes_prev = lag(tonnes), delta_t = tonnes - lag(tonnes)) |>
  filter(!is.na(delta_t)) |>
  head(6)
```

## Q11 · if_else() & case_when() {#q11}
```{r q11}
fishery_bins <- fishery |>
  mutate(
    size_bin = case_when(
      tonnes < 30 ~ "small", # first matching rule
      tonnes < 70 ~ "medium",
      TRUE ~ "large" # catch-all required
    )
  )
count(fishery_bins, size_bin) # bin frequencies
```

**Ex 2.**

```{r q11-reef}
reef |>
  mutate(depth_zone = case_when(
    depth_m < 5 ~ "upper",
    depth_m < 10 ~ "mid",
    TRUE ~ "deep"
  )) |>
  count(site, depth_zone)
```

**Ex 2.**

```{r q11-acoustic}
acoustic |>
  mutate(active = if_else(detections >= 4, "high", "low")) |>
  count(reef_zone, active)
```

## Q12 · across() many columns {#q12}
```{r q12}
biolog_scaled <- biolog |> # burst time series
  mutate(across(c(depth_m, pitch_deg), ~ as.numeric(scale(.)))) # z-scores
summary(biolog_scaled$depth_m) # mean ~0, sd ~1
```

**Ex 2.**

```{r q12-reef}
reef |>
  filter(site == "North") |>
  mutate(across(c(count, depth_m), ~ as.numeric(scale(.)))) |>
  summarise(across(c(count, depth_m), mean))
```

**Ex 2.**

```{r q12-fishery}
fishery |>
  group_by(port) |>
  summarise(across(tonnes, list(mean = mean, sd = sd), .names = "{.fn}_{.col}"))
```

## Q13 · group_modify() custom tables {#q13}
```{r q13}
sp_range <- reef |> # species counts
  group_by(species) |>
  group_modify(~ {
    tibble( # return tibble shape
      min_n = min(.x$count),
      max_n = max(.x$count),
      range_n = max(.x$count) - min(.x$count)
    )
  })
sp_range
```

**Ex 2.**

```{r q13-fishery}
fishery |>
  group_by(port) |>
  group_modify(~ tibble(min_t = min(.x$tonnes), max_t = max(.x$tonnes)))
```

**Ex 2.**

```{r q13-acoustic}
acoustic |>
  group_by(animal_id) |>
  group_modify(~ {
    .x |>
      group_by(receiver) |>
      summarise(d = sum(detections), .groups = "drop") |>
      slice_max(d, n = 1)
  })
```

## Q14 · mutate(.by = ) {#q14}
```{r q14}
acoustic_cent <- acoustic |>
  mutate(
    det_cent = detections - mean(detections), # overall mean if no .by
    det_by_an = detections - mean(detections), .by = animal_id # within-tag mean
  )
head(acoustic_cent) # compare columns
```

## Q15 · left_join() metadata {#q15}
```{r q15}
ports <- tibble( # small lookup table
  port = c("Cairns", "Broome", "Mooloolaba"),
  region = c("GBR", "NW", "SEQ")
)
fishery_geo <- fishery |>
  left_join(ports, by = "port")
count(fishery_geo, region)
```

**Ex 2.**

```{r q15-acoustic}
tag_meta <- tibble(
  animal_id = c("A101", "A102", "A103"),
  tag_type = c("V16", "V16", "V13")
)
acoustic |>
  left_join(tag_meta, by = "animal_id") |>
  count(tag_type, reef_zone)
```

**Ex 2.**

```{r q15-reef}
site_meta <- tibble(site = c("North", "South", "East"), zone = c("MPA", "Open", "MPA"))
reef |>
  left_join(site_meta, by = "site") |>
  distinct(site, zone)
```

## Q16 · bind_rows() stack surveys {#q16}
```{r q16}
reef_a <- reef |> filter(site == "North") # first leg
reef_b <- reef |> filter(site == "South") |> # second leg
  mutate(transect_id = paste0(transect_id, "_b"))
stacked <- bind_rows(
  mutate(reef_a, leg = "north"),
  mutate(reef_b, leg = "south")
) # tagged stack
nrow(stacked)
```

## Q17 · pivot_longer() {#q17}
```{r q17}
wide_demo <- fishery |> # long landings
  filter(year == 2020) |>
  select(species, tonnes) |> # minimal columns
  pivot_wider(names_from = species, values_from = tonnes) # wide for exercise
long_demo <- wide_demo |> # now melt back
  pivot_longer(everything(), names_to = "species", values_to = "tonnes", values_drop_na = TRUE)
long_demo # long form restored
```

**Ex 2.**

```{r q17-reef}
reef |>
  filter(transect_id == "T01") |>
  select(species, count) |>
  pivot_wider(names_from = species, values_from = count) |>
  pivot_longer(everything(), names_to = "species", values_to = "count")
```

**Ex 2.**

```{r q17-biolog}
biolog |>
  filter(burst_id == 1) |>
  select(seconds, depth_m, pitch_deg) |>
  pivot_longer(c(depth_m, pitch_deg), names_to = "sensor", values_to = "value") |>
  head(6)
```

## Q18 · pivot_wider() {#q18}
```{r q18}
ac_wide <- acoustic |> # long detections
  group_by(animal_id, receiver) |> # unique pairs
  summarise(dets = sum(detections), .groups = "drop") |> # aggregate duplicates
  pivot_wider(names_from = receiver, values_from = dets, values_fill = 0)
ac_wide # wide receiver columns
```

## Q19 · replace_na() & coalesce() {#q19}
```{r q19}
reef_na <- reef |> # original reef
  mutate(depth_m = if_else(species == "Grouper", NA_real_, depth_m)) # inject NA demo
reef_fix <- reef_na |> # repair for plotting
  mutate(depth_m = replace_na(depth_m, median(reef$depth_m, na.rm = TRUE)))
sum(is.na(reef_fix$depth_m)) # should be zero
```

## Q20 · dplyr capstone pipeline {#q20}
```{r q20}
cap <- fishery |> # start landings
  filter(year >= 2019) |> # recent years
  group_by(species, port) |> # reporting cells
  summarise(total_t = sum(tonnes), .groups = "drop") |>
  arrange(desc(total_t))
ggplot(cap, aes(x = reorder(species, total_t), y = total_t, fill = port)) +
  geom_col(position = "dodge") +
  coord_flip() + # readable labels
  labs(title = "Landings by species and port (2019+)", x = NULL, y = "Tonnes")
```

