---
title: "Functional programming · Beginner (Base R)"
description: "Beginner · base R"
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Functional programming · Beginner (Base R)
Base R: vectors, `[`, `apply`, loops. ~10 min/quest; harder quests show 2–3 datasets.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
```

## Q1 · Vectors, names, and types {#q01}
```{r q01}
counts <- reef$count
sites <- reef$site # site names
names(counts) <- NULL
length(counts)
typeof(counts) # internal storage type
unique(sites)
```

## Q2 · Subsetting `[`, `$`, `[[` {#q02}
```{r q02}
recent <- fishery[fishery$year >= 2020, ]
tons_snapper <- fishery$tonnes[fishery$species == "Snapper"] # vector of Snapper landings
cols <- fishery[, c("year", "species", "tonnes")]
first_port <- fishery[["port"]][1]
nrow(recent)
```

## Q3 · if, else, and ifelse {#q03}
```{r q03}
high <- acoustic$detections > 4
flag <- ifelse(high, "revisit", "ok")
table(flag)
if (any(is.na(acoustic$detections))) {
  message("Found missing detection counts")
} else {
  message("No NA in detections")
}
```

**Ex 2.**

```{r q03-reef}
reef$depth_class <- ifelse(reef$depth_m < 6, "shallow", "deep")
table(reef$depth_class)
```

**Ex 2.**

```{r q03-fishery}
fishery$big_trip <- ifelse(fishery$tonnes >= 50, "major", "routine")
if (any(fishery$tonnes > 100, na.rm = TRUE)) {
  message("At least one landing exceeds 100 t — check port log")
}
table(fishery$big_trip)
```

## Q4 · for loops on real rows {#q04}
```{r q04}
burst_ids <- unique(biolog$burst_id)
mean_depths <- numeric(length(burst_ids))
for (i in seq_along(burst_ids)) {
  bid <- burst_ids[i]
  rows <- biolog$depth_m[biolog$burst_id == bid]
  mean_depths[i] <- mean(rows)
}
names(mean_depths) <- paste0("burst_", burst_ids)
mean_depths
```

**Ex 2.**

```{r q04-fishery}
ports <- unique(fishery$port) # each landing port once
port_tot <- numeric(length(ports)) # pre-allocate totals
for (j in seq_along(ports)) {
  p <- ports[j]
  port_tot[j] <- sum(fishery$tonnes[fishery$port == p])
}
names(port_tot) <- ports
port_tot
```

**Ex 2.**

```{r q04-acoustic}
tags <- unique(acoustic$animal_id) # tagged animals in file
tag_sum <- numeric(length(tags)) # one total per tag
for (k in seq_along(tags)) {
  tag_sum[k] <- sum(acoustic$detections[acoustic$animal_id == tags[k]])
}
names(tag_sum) <- tags
tag_sum
```

## Q5 · Writing functions {#q05}
```{r q05}
tonnes_to_kg <- function(tonnes, factor = 1000) { # factor default matches metric
  tonnes * factor
}
fishery$kg <- tonnes_to_kg(fishery$tonnes) # apply to landing weights
sum(fishery$kg, na.rm = TRUE) # total kg landed in file
tonnes_to_kg(1.5) # scalar check: expect 1500
```

## Q6 · apply() on fisheries matrix {#q06}
```{r q06}
sp <- unique(fishery$species) # row labels for matrix
pr <- unique(fishery$port) # column labels
mat <- matrix(0, nrow = length(sp), ncol = length(pr)) # pre-fill with zeros
rownames(mat) <- sp # name rows for readable output
colnames(mat) <- pr # name columns
for (i in seq_along(sp)) {
  for (j in seq_along(pr)) {
    w <- fishery$tonnes[fishery$species == sp[i] & fishery$port == pr[j]]
    mat[i, j] <- if (length(w)) mean(w) else NA_real_
  }
}
apply(mat, 2, mean, na.rm = TRUE)
```

**Ex 2.**

```{r q06-reef}
reef_mat <- as.matrix(xtabs(count ~ transect_id + species, data = reef)) # wide count matrix
apply(reef_mat, 1, sum)[1:4] # total fish on first four
apply(reef_mat, 2, mean)
```

**Ex 2.**

```{r q06-acoustic}
ac_mat <- as.matrix(xtabs(detections ~ animal_id + reef_zone, data = acoustic))
apply(ac_mat, 2, sum) # detections summed per reef zone
```

## Q7 · lapply() lists {#q07}
```{r q07}
by_site <- split(reef, reef$site) # list of data frames by
richness <- lapply(by_site, function(df) { # function runs on each site
  length(unique(df$species)) # count distinct species names
})
richness # list of one integer per
unlist(richness)
```

**Ex 2.**

```{r q07-fishery}
by_year <- split(fishery$tonnes, fishery$year) # list of tonne vectors per
year_sums <- lapply(by_year, sum)
unlist(year_sums)
```

**Ex 2.**

```{r q07-biolog}
by_burst <- split(biolog$depth_m, biolog$burst_id)
max_depths <- lapply(by_burst, max)
head(unlist(max_depths))
```

## Q8 · sapply() simplification traps {#q08}
```{r q08}
by_animal <- split(acoustic$detections, acoustic$animal_id) # list of detection vectors
via_sapply <- sapply(by_animal, max) # simplified numeric vector if uniform
via_lapply <- lapply(by_animal, max) # always a list
str(via_sapply) # inspect simplified structure
str(via_lapply) # compare: list of scalars
```

## Q9 · vapply() safe types {#q09}
```{r q09}
by_receiver <- split(acoustic$detections, acoustic$receiver) # group detections
totals <- vapply( # vapply needs FUN.VALUE prototype
  by_receiver,
  sum,
  FUN.VALUE = numeric(1) # each element must be length-1
)
sort(totals, decreasing = TRUE) # which receivers saw most pings
```

## Q10 · split() + lapply() {#q10}
```{r q10}
counts_by_sp <- split(reef$count, reef$species) # list of count vectors per
mean_counts <- lapply(counts_by_sp, mean)
as.data.frame(unlist(mean_counts)) # quick rectangular view
# named vector of means:
sapply(counts_by_sp, mean)
```

**Ex 2.**

```{r q10-fishery}
tonnes_by_sp <- split(fishery$tonnes, fishery$species)
lapply(tonnes_by_sp, mean)
```

**Ex 2.**

```{r q10-acoustic}
by_rx <- split(acoustic$detections, acoustic$receiver)
sapply(by_rx, sum) # named vector of station totals
```

## Q11 · tapply() group means {#q11}
```{r q11}
tapply( # classic split-apply-combine
  fishery$tonnes, # numeric vector to summarise
  fishery$species,
  mean # function applied per group
)
```

## Q12 · aggregate() classic {#q12}
```{r q12}
agg_zone <- aggregate( # formula: response ~ grouping
  detections ~ reef_zone, # sum detections per zone
  data = acoustic,
  FUN = sum
)
agg_zone # one row per zone
```

## Q13 · mapply() paired vectors {#q13}
```{r q13}
b1 <- biolog[biolog$burst_id == 1, ]
pitch <- b1$pitch_deg # pitch vector in time order
secs <- b1$seconds
d_pitch <- mapply(function(a, b) abs(b - a),
                  pitch[-length(pitch)], # earlier second
                  pitch[-1]) # next second
head(d_pitch) # first few step changes
```

**Ex 2.**

```{r q13-fishery}
ord <- order(fishery$year) # sort by year for pairing
t <- fishery$tonnes[ord] # tonnes in year order
increase <- mapply(function(prev, curr) curr > prev, t[-length(t)], t[-1]) # TRUE if this trip >
head(increase)
```

**Ex 2.**

```{r q13-reef}
mapply(function(c, d) if (d > 0) c / d else NA_real_, reef$count, reef$depth_m)[1:6]
```

## Q14 · set.seed() & sample() {#q14}
```{r q14}
set.seed(20260721) # lock RNG for this quest
ids <- unique(reef$transect_id) # all transect labels
audit_ids <- sample(ids, size = 4) # pick 4 transects without replacement
reef[reef$transect_id %in% audit_ids, ] # rows selected for audit
```

## Q15 · order() and sorting logic {#q15}
```{r q15}
ord <- order(fishery$year, -fishery$tonnes) # year ascending, tonnes descending
fishery_sorted <- fishery[ord, ] # reorder all columns together
head(fishery_sorted, 8)
```

## Q16 · unique() & table() {#q16}
```{r q16}
unique(acoustic$animal_id) # how many tagged animals
table(acoustic$reef_zone)
table(acoustic$animal_id, acoustic$reef_zone) # two-way contingency
```

## Q17 · str(), class(), debugging {#q17}
```{r q17}
str(biolog) # column types and row count
class(biolog) # should be data.frame
sapply(biolog, class)
summary(biolog$depth_m)
```

## Q18 · Nested if-else for QC {#q18}
```{r q18}
qc <- character(nrow(reef))
for (i in seq_len(nrow(reef))) { # row-wise loop for clarity
  if (is.na(reef$depth_m[i])) { # missing depth first priority
    qc[i] <- "missing_depth"
  } else if (reef$depth_m[i] < 0) { # impossible depth
    qc[i] <- "bad_depth"
  } else if (reef$count[i] == 0) { # true zero count
    qc[i] <- "zero_count"
  } else {
    qc[i] <- "ok"
  }
}
table(qc) # summary of flags
```

**Ex 2.**

```{r q18-fishery}
fqc <- rep("ok", nrow(fishery))
for (i in seq_len(nrow(fishery))) {
  if (is.na(fishery$tonnes[i])) {
    fqc[i] <- "missing_tonnes"
  } else if (fishery$port[i] == "Broome" && fishery$tonnes[i] > 60) {
    fqc[i] <- "review_broome"
  } else if (fishery$tonnes[i] == 0) {
    fqc[i] <- "zero_landings"
  }
}
table(fqc)
```

**Ex 2.**

```{r q18-acoustic}
ac_qc <- ifelse(is.na(acoustic$detections), "missing",
  ifelse(acoustic$reef_zone == "Lagoon" & acoustic$detections > 5, "lagoon_hot",
    ifelse(acoustic$detections == 0, "zero", "ok")))
table(ac_qc)
```

## Q19 · Loop vs vectorised choice {#q19}
```{r q19}
# vectorised approach using ave() — mean pitch per burst_id
centred <- biolog$pitch_deg - ave(biolog$pitch_deg, biolog$burst_id, FUN = mean)
# loop approach for same result (first 50 rows demo)
loop_cent <- biolog$pitch_deg
for (i in 1:50) {
  bid <- biolog$burst_id[i]
  loop_cent[i] <- biolog$pitch_deg[i] - mean(biolog$pitch_deg[biolog$burst_id == bid])
}
all.equal(loop_cent, centred[1:50]) # methods should agree
```

**Ex 2.**

```{r q19-reef}
vec_bio <- reef$count * 0.1 # vectorised
loop_bio <- numeric(20)
for (i in 1:20) { loop_bio[i] <- reef$count[i] * 0.1 }
all.equal(vec_bio[1:20], loop_bio)
```

**Ex 2.**

```{r q19-fishery}
ord <- order(fishery$year, fishery$species)
t <- fishery$tonnes[ord]
all.equal(cumsum(t), { out <- numeric(length(t)); for (i in seq_along(t)) out[i] <- sum(t[1:i]); out })
```

## Q20 · Base R mini-capstone {#q20}
```{r q20}
library(ggplot2) # only for quick plot at
ac <- acoustic # working copy
ac$hot <- ac$detections >= 5
by_an <- aggregate(detections ~ animal_id, data = ac, FUN = sum) # totals per tag
zone <- aggregate(reef_zone ~ animal_id, data = ac, FUN = function(x) unique(x)[1])
merged <- merge(by_an, zone, by = "animal_id")
ggplot(merged, aes(x = reorder(animal_id, detections), y = detections, fill = reef_zone)) +
  geom_col() + # bar chart of totals
  coord_flip() + # readable animal labels
  labs(title = "Detections per tagged animal", x = NULL, y = "Total detections")
```

