---
title: "Level 1 · Base R"
description: "Twenty quests — vectors, if/else, loops, apply, lapply, split-apply in base R."
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Level 1 · Base R

Marine field data arrive as messy tables long before anyone opens a modelling package. Base R is the grammar underneath **dplyr**, **data.table**, and most AI-generated snippets — if you cannot read vectors, indices, and `apply` family calls, you cannot spot when autocomplete rewired your grouping variable.

This level follows how people actually learn technical skills: connect new ideas to what you already know, practice in short segments, and get quick feedback ([Bransford, Brown & Cocking, 2000](https://nap.nationalacademies.org/catalog/9853/how-people-learn-brain-mind-experience-and-school-expanded-edition) — *How People Learn*). Scientific coding is also hypothesis testing in miniature: you predict what a function should return, run it, and revise ([Klahr & Dunbar, 1988](https://doi.org/10.1016/0166-4115(88)90004-2)). Treat each ~10-minute quest as a controlled experiment on real reef, fishery, acoustic, and biologging tables — not as trivia to memorise.

Use AI as a fast typist **after** you can narrate each line aloud. When a chatbot suggests `sapply()` where you needed `vapply()`, your mental model of return types is what keeps the manuscript honest.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width=7, fig.height=4.2, dpi=120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
biolog <- read.csv(data_path("biologging_burst.csv"))
```

## Q1 · Vectors, names, and types {#q01}

Ecological summaries start as **atomic vectors** — counts per species, depths, years — before they become data frames. R keeps a `mode` and optional `names()` so you can label transects without a spreadsheet column yet.

Building a named vector trains the same habit you need when checking AI output: inspect structure before chaining functions. If types are wrong (character where you expected numeric), downstream code fails in confusing places.

We pull `count` from the reef transect file into a plain vector so you see what lives underneath `$` and `[` later.

```{r q01}
counts <- reef$count          # extract the count column as a numeric vector
sites <- reef$site              # character vector of site names
names(counts) <- NULL         # ensure no stray names on counts for now
length(counts)                  # how many species rows in the file
typeof(counts)                  # internal storage type (double for these counts)
unique(sites)                 # which reef sites appear in the survey
```

## Q2 · Subsetting `[`, `$`, `[[` {#q02}

Subsetting is how you **focus** on one transect, one species, or one column without copying the whole dataset. Base R offers three related operators; they look interchangeable until you subset lists or programmatic indices.

On data frames, `$` is readable for fixed column names; `[` handles rows and columns together; `[[` drops a dimension and is essential for list columns later. AI often picks the prettiest form — you verify it matches the shape you need.

Practice on `fishery` landings: keep recent years and selected columns.

```{r q02}
recent <- fishery[fishery$year >= 2020, ]           # rows where year is 2020 or later
tons_snapper <- fishery$tonnes[fishery$species == "Snapper"]  # vector of Snapper landings
cols <- fishery[, c("year", "species", "tonnes")]      # two-dimensional subset (data frame)
first_port <- fishery[["port"]][1]                      # [[ extracts column then first value
nrow(recent)                                            # how many landing records matched
```

## Q3 · if, else, and ifelse {#q03}

Field QC rules are conditional: **if** depth is impossible, flag the row; **else if** count is zero, mark as absent; **else** keep for analysis. Base R expresses that with `if`/`else` for single decisions and `ifelse()` for vectorised branches.

Vectorised `ifelse()` is what AI loves to drop into pipelines — but it evaluates both branches for every element, which bites when branches call slow functions. Knowing when to use scalar `if` trains careful thinking about data length.

Flag acoustic detections with high counts as candidate revisits.

```{r q03}
high <- acoustic$detections > 4                    # logical vector: TRUE where detections exceed 4
flag <- ifelse(high, "revisit", "ok")                   # same-length character vector of flags
table(flag)                                             # tally how many rows in each category
if (any(is.na(acoustic$detections))) {                # scalar if: only runs when NA present
  message("Found missing detection counts")             # side effect for QC log
} else {
  message("No NA in detections")                        # alternate branch when data clean
}
```

## Q4 · for loops on real rows {#q04}

A **`for` loop** repeats code for each index — useful when row logic is sequential (running totals, burst windows) or too awkward to vectorise on day one. Loops are slower on huge data but **transparent**: you can `print()` inside and watch state change.

Students often let AI replace loops with `apply` prematurely. Keeping one explicit loop in your toolkit helps you debug what the vectorised version should do.

Sum biologging depth within each burst id by looping over unique bursts.

```{r q04}
burst_ids <- unique(biolog$burst_id)                # which dive bursts exist
mean_depths <- numeric(length(burst_ids))               # pre-allocate results vector
for (i in seq_along(burst_ids)) {                       # i indexes positions in burst_ids
  bid <- burst_ids[i]                                   # current burst identifier
  rows <- biolog$depth_m[biolog$burst_id == bid]        # depths belonging to this burst
  mean_depths[i] <- mean(rows)                          # store mean for this burst
}
names(mean_depths) <- paste0("burst_", burst_ids)       # label output for inspection
mean_depths                                             # print means per burst
```

## Q5 · Writing functions {#q05}

A **function** packages a reusable computation with named arguments — the same abstraction AI invokes when it writes `my_helper <- function(x) ...`. Writing one yourself forces you to decide inputs, defaults, and return value.

Good functions are small and testable on one row before scaling. That mirrors Klahr–Dunbar-style testing: one clear prediction, one run, revise.

Encapsulate tonnes-to-kilograms conversion for fishery landings.

```{r q05}
tonnes_to_kg <- function(tonnes, factor = 1000) {   # factor default matches metric conversion
  tonnes * factor                                      # vectorised multiply works on whole column
}
fishery$kg <- tonnes_to_kg(fishery$tonnes)              # apply to landing weights
sum(fishery$kg, na.rm = TRUE)                           # total kg landed in file
tonnes_to_kg(1.5)                                       # scalar check: expect 1500
```

## Q6 · apply() on fisheries matrix {#q06}

`apply()` moves a function across **margins** of an array — typically rows or columns of a matrix. It is the classic base R substitute for grouped summaries before **dplyr**.

The margin argument `1` means “each row”, `2` means “each column”. AI frequently swaps them; your job is to picture the matrix orientation.

Build a species-by-port matrix of mean tonnes and take column means.

```{r q06}
sp <- unique(fishery$species)                       # row labels for matrix
pr <- unique(fishery$port)                              # column labels
mat <- matrix(0, nrow = length(sp), ncol = length(pr))  # pre-fill with zeros
rownames(mat) <- sp                                     # name rows for readable output
colnames(mat) <- pr                                     # name columns
for (i in seq_along(sp)) {                              # fill cells with mean tonnes
  for (j in seq_along(pr)) {
    w <- fishery$tonnes[fishery$species == sp[i] & fishery$port == pr[j]]
    mat[i, j] <- if (length(w)) mean(w) else NA_real_
  }
}
apply(mat, 2, mean, na.rm = TRUE)                       # mean down rows: one value per port
```

## Q7 · lapply() lists {#q07}

`lapply()` applies a function to each element of a **list** and always returns a list. That predictability makes it safer than `sapply()` when teaching — and when auditing AI code.

Splitting data by site naturally yields a list of mini data frames; `lapply()` is how you run the same summary on each piece.

Split reef counts by site and compute species richness (unique species per site).

```{r q07}
by_site <- split(reef, reef$site)                    # list of data frames by site
richness <- lapply(by_site, function(df) {              # function runs on each site chunk
  length(unique(df$species))                            # count distinct species names
})
richness                                              # list of one integer per site
unlist(richness)                                        # flatten to named vector
```

## Q8 · sapply() simplification traps {#q08}

`sapply()` tries to **simplify** `lapply()` output to a vector or matrix when shapes match. Convenient — until one site returns a different length and R silently returns a list or a matrix you did not expect.

When AI uses `sapply()`, ask: “What if one element is length 0?” Training on traps builds skepticism worth more than memorising syntax.

Compute max detections per animal; compare simplified type to `lapply`.

```{r q08}
by_animal <- split(acoustic$detections, acoustic$animal_id)  # list of detection vectors
via_sapply <- sapply(by_animal, max)                    # simplified numeric vector if uniform
via_lapply <- lapply(by_animal, max)                    # always a list
str(via_sapply)                                         # inspect simplified structure
str(via_lapply)                                         # compare: list of scalars
```

## Q9 · vapply() safe types {#q09}

`vapply()` is like `sapply()` but you declare **`FUN.VALUE`** — a template of the return type and length. If the function deviates, R errors immediately instead of coercing quietly.

That explicit contract is perfect QC culture for ecological scripts: no silent recycling of wrong-length vectors into a matrix.

Return numeric total detections per receiver with a enforced scalar template.

```{r q09}
by_receiver <- split(acoustic$detections, acoustic$receiver)   # group detections
totals <- vapply(                                       # vapply needs FUN.VALUE prototype
  by_receiver,
  sum,
  FUN.VALUE = numeric(1)                                # each element must be length-1 numeric
)
sort(totals, decreasing = TRUE)                         # which receivers saw most pings
```

## Q10 · split() + lapply() {#q10}

**`split()`** cuts a vector (or column) by a factor-like grouping variable; paired with **`lapply()`** it is the base R ancestor of `group_by()` + `summarise()`.

Understanding this pair helps you read older ecology code and translate AI suggestions into equivalent base R when packages are unavailable.

Split reef `count` by `species` and get mean abundance per species.

```{r q10}
counts_by_sp <- split(reef$count, reef$species)     # list of count vectors per species
mean_counts <- lapply(counts_by_sp, mean)               # mean within each species
as.data.frame(unlist(mean_counts))                     # quick rectangular view
# named vector of means:
sapply(counts_by_sp, mean)
```

## Q11 · tapply() group means {#q11}

`tapply()` applies a function to subsets defined by a **factor-like index** and returns a compact array — often the fastest mental path to “mean depth by site”.

Arguments are `tapply(X, INDEX, FUN)`. AI sometimes nests `tapply` inside `tapply`; knowing the flat form keeps you grounded.

Mean landing tonnes by species using fishery data.

```{r q11}
tapply(                                               # classic split-apply-combine
  fishery$tonnes,                                       # numeric vector to summarise
  fishery$species,                                      # grouping index
  mean                                                  # function applied per group
)
```

## Q12 · aggregate() classic {#q12}

`aggregate()` produces **grouped summaries as a data frame**, friendlier for merging than raw `tapply` arrays. Formula interface `y ~ x` reads like statistical notation.

It is still base R — no **dplyr** required — and appears in legacy fisheries reports you may need to reproduce.

Total detections by reef zone from acoustic telemetry.

```{r q12}
agg_zone <- aggregate(                              # formula: response ~ grouping
  detections ~ reef_zone,                               # sum detections per zone
  data = acoustic,
  FUN = sum
)
agg_zone                                                # one row per zone
```

## Q13 · mapply() paired vectors {#q13}

`mapply()` is **multivariate `apply`**: pass several vectors of equal length, and the function receives the i-th element from each. Think paired depth and pitch rows within a burst.

It generalises to `Map()` in base R and `map2()` in **purrr** later — same mental picture.

Compute absolute pitch change between consecutive seconds inside burst 1.

```{r q13}
b1 <- biolog[biolog$burst_id == 1, ]               # restrict to one burst
pitch <- b1$pitch_deg                                   # pitch vector in time order
secs <- b1$seconds                                      # matching time index
d_pitch <- mapply(function(a, b) abs(b - a),            # pairwise function
                  pitch[-length(pitch)],                 # earlier second
                  pitch[-1])                             # next second
head(d_pitch)                                           # first few step changes
```

## Q14 · set.seed() & sample() {#q14}

Reproducible ecology needs **`set.seed()`** before **`sample()`** so subsampling transects or simulating null models can be re-run by reviewers — and matched when you paste reprexes into AI chats.

Sampling without a fixed seed is not wrong for field work, but wrong for teaching and debugging.

Draw a reproducible random subset of transect ids for double-checking counts.

```{r q14}
set.seed(20260721)                                  # lock RNG for this quest
ids <- unique(reef$transect_id)                         # all transect labels
audit_ids <- sample(ids, size = 4)                      # pick 4 transects without replacement
reef[reef$transect_id %in% audit_ids, ]                 # rows selected for audit
```

## Q15 · order() and sorting logic {#q15}

Sorting in R is usually **`order()`**, not `sort()` on the whole table. `order()` returns **indices** you use to reorder multiple columns consistently — critical when AI sorts one column but leaves others aligned wrong.

Negative indices in `order()` flip descending sort.

Sort fishery rows by tonnes within year.

```{r q15}
ord <- order(fishery$year, -fishery$tonnes)        # year ascending, tonnes descending
fishery_sorted <- fishery[ord, ]                        # reorder all columns together
head(fishery_sorted, 8)                                 # inspect top rows
```

## Q16 · unique() & table() {#q16}

`unique()` lists distinct values; **`table()`** cross-tabulates frequencies — the fastest check on whether AI duplicated join keys or miscoded species names.

Always run `table()` on grouping columns before trusting grouped summaries.

Tabulate animals and receivers in the acoustic file.

```{r q16}
unique(acoustic$animal_id)                          # how many tagged animals
table(acoustic$reef_zone)                               # detections rows per zone label
table(acoustic$animal_id, acoustic$reef_zone)           # two-way contingency
```

## Q17 · str(), class(), debugging {#q17}

**`str()`** compactly shows structure; **`class()`** and **`typeof()`** distinguish data.frame vs tibble vs list. These are the first commands to run when AI output “looks wrong” but does not error.

Debugging is hypothesis revision — predict structure, compare to `str()`, adjust.

Inspect biologging columns before computing summaries.

```{r q17}
str(biolog)                                           # column types and row count
class(biolog)                                           # should be data.frame
sapply(biolog, class)                                   # class of each column
summary(biolog$depth_m)                                 # quick numeric QC
```

## Q18 · Nested if-else for QC {#q18}

Nested **`if`/`else`** handles **priority rules** where only one label should apply — e.g., missing depth overrides high-count flags. Nested logic mirrors decision trees in field protocols.

Keep branches shallow; if nesting exceeds three levels, consider a lookup table — but nested `if` remains the readable skeleton.

Assign QC labels to reef rows by depth and count.

```{r q18}
qc <- character(nrow(reef))                         # pre-allocate label vector
for (i in seq_len(nrow(reef))) {                        # row-wise loop for clarity
  if (is.na(reef$depth_m[i])) {                         # missing depth first priority
    qc[i] <- "missing_depth"
  } else if (reef$depth_m[i] < 0) {                     # impossible depth
    qc[i] <- "bad_depth"
  } else if (reef$count[i] == 0) {                     # true zero count
    qc[i] <- "zero_count"
  } else {
    qc[i] <- "ok"                                       # passes basic QC
  }
}
table(qc)                                               # summary of flags
```

## Q19 · Loop vs vectorised choice {#q19}

Vectorised operations let R push work to compiled C loops — faster and often clearer for element-wise math. **`for` loops** still win for complex state, irregular sequences, or teaching.

Compare both on the same task and time only when data are large; for small reef tables, readability dominates.

Centre pitch within each burst: vectorised subtract vs loop.

```{r q19}
# vectorised approach using ave() — mean pitch per burst_id
centred <- biolog$pitch_deg - ave(biolog$pitch_deg, biolog$burst_id, FUN = mean)
# loop approach for same result (first 50 rows demo)
loop_cent <- biolog$pitch_deg
for (i in 1:50) {
  bid <- biolog$burst_id[i]
  loop_cent[i] <- biolog$pitch_deg[i] - mean(biolog$pitch_deg[biolog$burst_id == bid])
}
all.equal(loop_cent, centred[1:50])                     # methods should agree
```

## Q20 · Base R mini-capstone {#q20}

Capstone quests integrate split-apply-combine, functions, and QC without packages. This mirrors a real morning: import CSV, flag issues, summarise by stratum, export a small table for your PI.

Before asking AI to “clean this dataset”, run this checklist yourself — then use AI for boilerplate you can verify.

Pipeline: flag acoustic rows, aggregate by animal, join unique zones, plot top totals.

```{r q20}
library(ggplot2)                                      # only for quick plot at end
ac <- acoustic                                          # working copy
ac$hot <- ac$detections >= 5                            # logical hot-spot flag
by_an <- aggregate(detections ~ animal_id, data = ac, FUN = sum)  # totals per tag
zone <- aggregate(reef_zone ~ animal_id, data = ac, FUN = function(x) unique(x)[1])
merged <- merge(by_an, zone, by = "animal_id")          # base R join
ggplot(merged, aes(x = reorder(animal_id, detections), y = detections, fill = reef_zone)) +
  geom_col() +                                          # bar chart of totals
  coord_flip() +                                        # readable animal labels
  labs(title = "Detections per tagged animal", x = NULL, y = "Total detections")
```

