---
title: "Computer science crash course"
description: "CS crash course"
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Computer science crash course for ecologists
Types, memory, debugging, reproducibility. Light R.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.2, dpi = 120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
```

## Q1 · Programs, data, and R {#q01}
```{r q01}
input_rows <- nrow(reef)
output_sites <- length(unique(reef$site)) # how many sites after summarising
c(input_rows = input_rows, output_sites = output_sites)
```

## Q2 · Assignment and names {#q02}
```{r q02}
mean_depth <- mean(reef$depth_m) # store summary under a clear
mean_depth
rm(mean_depth) # remove binding
```

## Q3 · Types and coercion {#q03}
```{r q03}
typeof(reef$count) # should be double/integer
char_counts <- as.character(reef$count[1:3]) # simulate import as text
as.numeric(char_counts) # explicit conversion back
```

## Q4 · Memory: copy vs modify {#q04}
```{r q04}
a <- reef$count
b <- a
b[1] <- 999L # replace triggers copy for b
a[1] # original column unchanged in reef
```

## Q5 · Files and paths {#q05}
```{r q05}
getwd() # current working directory
file.path("ai-tutorials", "data", "reef_transects.csv") # portable path string
file.exists(data_path("reef_transects.csv")) # TRUE when CSV is found
```

## Q6 · Functions as contracts {#q06}
```{r q06}
cpue <- function(catch, effort) { # catch / effort with guard
  if (effort <= 0) return(NA_real_) # invalid effort → missing
  catch / effort
}
cpue(catch = 40, effort = 8) # example call
```

## Q7 · Boolean logic {#q07}
```{r q07}
bad_depth <- reef$depth_m < 0
any(bad_depth, na.rm = TRUE) # any impossible depths?
all(!duplicated(acoustic$animal_id) | TRUE) # tags can repeat rows; logic
table(acoustic$detections > 3) # TRUE/FALSE counts
```

## Q8 · Algorithms: find and sort {#q08}
```{r q08}
i_max <- which.max(fishery$tonnes)
fishery[i_max, c("year", "species", "tonnes")] # inspect that record
order(fishery$tonnes, decreasing = TRUE)[1:3] # top three indices
```

## Q9 · Big-O intuition for n rows {#q09}
```{r q09}
n <- nrow(acoustic) # n for this file
c(rows = n, double_pass = 2 * n, nested_demo = length(unique(acoustic$animal_id)) * 10)
```

## Q10 · Vectors, lists, tables {#q10}
```{r q10}
str(reef$count)
str(split(reef$count, reef$site)) # list of vectors
str(reef) # data.frame of columns
```

## Q11 · Encoding and factors {#q11}
```{r q11}
f <- as.factor(reef$site) # site as categorical
levels(f) # allowed labels
as.integer(f)[1:5] # internal level codes
```

## Q12 · Reproducibility & seeds {#q12}
```{r q12}
set.seed(42) # lock RNG
sample(reef$transect_id, 3) # three random transects
set.seed(42)
sample(reef$transect_id, 3) # identical draw
```

## Q13 · Debugging strategies {#q13}
```{r q13}
str(acoustic)
summary(acoustic$detections) # range and NA check
stopifnot(nrow(acoustic) > 0) # assert expectation early
```

## Q14 · Version control mindset {#q14}
```{r q15}
z_score <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE) # reusable
z_score(reef$count)[1:5]
z_score(fishery$tonnes)[1:5]
```

## Q16 · Validate inputs early {#q16}
```{r q16}
need <- c("animal_id", "detections") # required columns
if (!all(need %in% names(acoustic))) stop("Missing columns: ", paste(setdiff(need, names(acoustic)), collapse = ", "))
if (any(acoustic$detections < 0, na.rm = TRUE)) stop("Negative detections found")
message("Input checks passed")
```

## Q17 · Composition and pipes {#q17}
```{r q17}
suppressPackageStartupMessages(library(dplyr))
reef |>
  filter(count > 0) |>
  group_by(site) |>
  summarise(total = sum(count), .groups = "drop")
```

## Q18 · Reading error messages {#q18}
```{r q18}
# Example: intentional typo (run to see error in console when knitting locally)
# mean(reef$counts) # counts vs count
tryCatch(mean(reef$counts), error = function(e) conditionMessage(e))
```

## Q19 · Documentation & help() {#q19}
```{r q19}
args(mean) # formal arguments
# help(mean) # run in RStudio for full help page
```

## Q20 · CS capstone plan {#q20}
```{r q20}
plan <- list(
  input = "acoustic_detections.csv",
  output = "detections per animal_id",
  checks = c("no negative detections", "animal_id not empty"),
  steps = c("read.csv", "group_by + summarise", "ggplot if needed")
)
plan
```

{{< include ../_includes/learn-more-cs.md >}}