---
title: "Computer science crash course"
description: "Twenty quests — CS principles for R ecologists: types, memory, algorithms, debugging, reproducibility."
date: 2026-07-21
sidebar: ai-curriculum
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: false
freeze: true
execute:
  warning: false
  message: false
---

## Computer science crash course for ecologists

You do not need a computer science degree to analyse reef counts or tag data — but a **thin layer of CS thinking** makes everything else click: why `factor` levels matter, why joins duplicate rows, why a loop is slow on a million detections, why your script works on your laptop but not on a collaborator’s.

These quests use the same rhythm as the functional programming tracks (~10 minutes, annotated R where it helps, reef / fishery / acoustic / biologging examples). The goal is **competence**: read code, write code, debug code — whether you type it yourself or use a tool later.

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.2, dpi = 120)
proj <- getwd()
if (!file.exists(file.path(proj, "ai-tutorials/data/reef_transects.csv"))) {
  proj <- normalizePath(file.path(proj, "..", ".."), mustWork = FALSE)
}
data_path <- function(name) file.path(proj, "ai-tutorials/data", name)
reef <- read.csv(data_path("reef_transects.csv"))
fishery <- read.csv(data_path("fishery_landings.csv"))
acoustic <- read.csv(data_path("acoustic_detections.csv"))
```

## Q1 · Programs, data, and R {#q01}

A **program** is instructions that transform **input data** into **output** (tables, plots, model fits). R is both a calculator and a language for chaining those transformations — your transect CSV is input; a species richness table is output.

Thinking in input → process → output matches how you design a field study: samples in, inference out.

```{r q01}
input_rows <- nrow(reef)                              # how many rows enter the pipeline
output_sites <- length(unique(reef$site))               # how many sites after summarising
c(input_rows = input_rows, output_sites = output_sites) # named result vector
```

## Q2 · Assignment and names {#q02}

**`<-`** stores a value under a **name** (object). Names should describe meaning (`mean_depth`), not typing order (`x1`). The environment is a lookup table from names to values — the same idea as labelled columns, but for interim results.

```{r q02}
mean_depth <- mean(reef$depth_m)                      # store summary under a clear name
mean_depth                                                # retrieve by name
rm(mean_depth)                                            # remove binding (free name)
```

## Q3 · Types and coercion {#q03}

Every value has a **type** (`numeric`, `character`, `logical`, …). **Coercion** converts types, sometimes silently — `"12"` + 1 fails; `as.numeric("12")` succeeds. Ecology CSVs often import counts as character if a stray `"NA"` or `"."` appears.

```{r q03}
typeof(reef$count)                                      # should be double/integer
char_counts <- as.character(reef$count[1:3])              # simulate import as text
as.numeric(char_counts)                                   # explicit conversion back
```

## Q4 · Memory: copy vs modify {#q04}

R often **shares** memory until you **modify** a copy (copy-on-write). Large acoustic tables feel slow when you copy them repeatedly in a loop. Prefer one pipeline that adds columns once, or tools (**data.table** `:=`) that modify in place when you mean to.

```{r q04}
a <- reef$count                                         # share column vector
b <- a                                                    # b points at same data until changed
b[1] <- 999L                                              # replace triggers copy for b
a[1]                                                      # original column unchanged in reef
```

## Q5 · Files and paths {#q05}

Scripts live in **directories**; data live in **files** with **paths**. **`getwd()`** is where R thinks you are; **`file.path()`** builds paths portably. Reproducible projects use relative paths from a project root — not `setwd("~/Desktop")` on one machine only.

```{r q05}
getwd()                                                 # current working directory
file.path("ai-tutorials", "data", "reef_transects.csv") # portable path string
file.exists(data_path("reef_transects.csv"))              # TRUE when CSV is found
```

## Q6 · Functions as contracts {#q06}

A **function** is a contract: **inputs** (arguments), **output** (return value), and **side effects** (files written, messages). Good contracts are small and testable — one function for “CPUE per trip”, not one script that does everything.

```{r q06}
cpue <- function(catch, effort) {                     # catch / effort with guard
  if (effort <= 0) return(NA_real_)                     # invalid effort → missing
  catch / effort
}
cpue(catch = 40, effort = 8)                              # example call
```

## Q7 · Boolean logic {#q07}

**Logical** values (`TRUE` / `FALSE`) drive **`if`**, **`filter()`**, and indexing. **`&`** is element-wise AND; **`&&`** is scalar AND (first pair only). **`any()`** / **`all()`** summarise vectors — “any depth negative?”, “all tags unique?”.

```{r q07}
bad_depth <- reef$depth_m < 0                           # logical vector
any(bad_depth, na.rm = TRUE)                              # any impossible depths?
all(!duplicated(acoustic$animal_id) | TRUE)               # tags can repeat rows; logic demo
table(acoustic$detections > 3)                              # TRUE/FALSE counts
```

## Q8 · Algorithms: find and sort {#q08}

An **algorithm** is a step-by-step recipe. **`which.max()`** finds an index; **`order()`** returns sort indices. For small ecology tables, “fast enough” beats clever — for million-row detections, algorithm choice matters (Quest 9).

```{r q08}
i_max <- which.max(fishery$tonnes)                      # row with largest landing
fishery[i_max, c("year", "species", "tonnes")]            # inspect that record
order(fishery$tonnes, decreasing = TRUE)[1:3]             # top three indices
```

## Q9 · Big-O intuition for n rows {#q09}

**Big-O** describes how work grows with **n** (rows). Doubling rows roughly doubles time for a single **`O(n)`** pass; nested loops over sites × species can be **`O(n²)`** on wide problems. Vectorised R is often **`O(n)`** in C — loops in pure R are fine for teaching, costly at IMOS scale.

```{r q09}
n <- nrow(acoustic)                                     # n for this file
c(rows = n, double_pass = 2 * n, nested_demo = length(unique(acoustic$animal_id)) * 10)
```

## Q10 · Vectors, lists, tables {#q10}

**Vector** = one type, one dimension. **List** = mixed objects. **Data frame** = table of columns (vectors of equal length). Ecology exports are tables; nested receiver lists are lists — know which you hold.

```{r q10}
str(reef$count)                                         # atomic vector
str(split(reef$count, reef$site))                       # list of vectors
str(reef)                                                 # data.frame of columns
```

## Q11 · Encoding and factors {#q11}

**Factors** store **levels** (categories) efficiently; **character** stores full strings. Stats models often want factors; joins want consistent spelling. **`forcats`** helps in tidy workflows; base **`as.factor()`** is enough to understand the idea.

```{r q11}
f <- as.factor(reef$site)                               # site as categorical
levels(f)                                                 # allowed labels
as.integer(f)[1:5]                                        # internal level codes
```

## Q12 · Reproducibility & seeds {#q12}

**Reproducibility** means someone else (or future you) gets the same result from the same code and data. **`set.seed()`** fixes pseudo-random draws for subsampling and simulation in papers and supplements.

```{r q12}
set.seed(42)                                            # lock RNG
sample(reef$transect_id, 3)                               # three random transects
set.seed(42)                                              # same seed again
sample(reef$transect_id, 3)                               # identical draw
```

## Q13 · Debugging strategies {#q13}

**Debugging** is hypothesis testing: predict, inspect, narrow. **`str()`**, **`head()`**, **`nrow()`**, **`summary()`**, and **`browser()`** (when needed) beat randomly changing code. Halve the pipeline until the bug appears — same as ruling out gear failure on a boat.

```{r q13}
str(acoustic)                                           # structure first
summary(acoustic$detections)                              # range and NA check
stopifnot(nrow(acoustic) > 0)                             # assert expectation early
```

## Q14 · Version control mindset {#q14}

**Version control** (Git) tracks **changes** to scripts over time — who changed what, when, and why. You do not need every Git command on day one; you need the habit: **small commits**, meaningful messages, one analysis per branch. Field notebooks have dates; code should too.

*(This quest is conceptual — no Git commands required in R.)*

## Q15 · DRY and small functions {#q15}

**DRY** (Don’t Repeat Yourself): one definition, many uses. Copy-pasted **`mutate()`** blocks drift apart; a function or shared script chunk stays consistent across reef and fishery reports.

```{r q15}
z_score <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)  # reusable
z_score(reef$count)[1:5]
z_score(fishery$tonnes)[1:5]
```

## Q16 · Validate inputs early {#q16}

**Fail fast**: check column names, types, and ranges at the top of a script. Clear **`stop()`** messages save hours when a CSV column was renamed (`Detections` vs `detections`).

```{r q16}
need <- c("animal_id", "detections")                   # required columns
if (!all(need %in% names(acoustic))) stop("Missing columns: ", paste(setdiff(need, names(acoustic)), collapse = ", "))
if (any(acoustic$detections < 0, na.rm = TRUE)) stop("Negative detections found")
message("Input checks passed")
```

## Q17 · Composition and pipes {#q17}

**Composition** chains small steps: output of step 1 feeds step 2. The **pipe** **`|>`** passes the left object as the first argument to the right — readable top-to-bottom narratives for dplyr pipelines.

```{r q17}
suppressPackageStartupMessages(library(dplyr))
reef |>
  filter(count > 0) |>
  group_by(site) |>
  summarise(total = sum(count), .groups = "drop")
```

## Q18 · Reading error messages {#q18}

Errors read **bottom-up**: what failed, then the call stack. **`object not found`** → typo in a name. **`non-numeric argument`** → wrong type. Save the full message when asking for help — it is data.

```{r q18}
# Example: intentional typo (run to see error in console when knitting locally)
# mean(reef$counts)  # counts vs count
tryCatch(mean(reef$counts), error = function(e) conditionMessage(e))
```

## Q19 · Documentation & help() {#q19}

**`?mean`**, **`help(mean)`**, and package vignettes are authoritative. **`args(mean)`** shows the argument list quickly. Writing a one-line comment per custom function is documentation your future self will thank you for.

```{r q19}
args(mean)                                              # formal arguments
# help(mean)                                            # run in RStudio for full help page
```

## Q20 · CS capstone plan {#q20}

Before writing code for a new dataset, sketch on paper:

1. **Inputs** — files, columns, units  
2. **Outputs** — tables, figures, model objects  
3. **Checks** — row counts, NA rules, keys  
4. **Steps** — import → clean → summarise → plot  

That plan is what good coursework, theses, and later **tool-assisted** coding all depend on — the tool fills syntax; you own the design.

```{r q20}
plan <- list(
  input = "acoustic_detections.csv",
  output = "detections per animal_id",
  checks = c("no negative detections", "animal_id not empty"),
  steps = c("read.csv", "group_by + summarise", "ggplot if needed")
)
plan
```
